---
title: "Expert Evaluation"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

# Data Processing

```{r, echo=FALSE}
library(tidyverse)
library(gtools)
library(psych)
library(irr)
library(knitr)

set.seed(8675309) # makes sure random numbers are reproducible
```

```{r}
options(readr.show_col_types = FALSE)
file_path <- "data/TPSai+-+Experts_15+April+2026_16.33.csv"
raw_data <- read_csv(file_path, n_max = 2)
colnames <- names(raw_data)
question_labels <- setNames(sub(" - .*", "", as.character(raw_data[1, ])), colnames)
raw_data <- read_csv(file_path, col_names = colnames, skip = 3)
```

```{r}
factors_map <- read_csv('data/factors.csv')
cols_tpsai <- factors_map$IID
choices <- c("Keep" = 1, "Drop" = 2, "Reword" = 3)
```

```{r}
# check that survey and factors correspond (should be empty dataframe)
data.frame(
  question_label = unlist(question_labels[cols_tpsai]),
  factor_item = factors_map$Item,
  match = unlist(question_labels[cols_tpsai]) == factors_map$Item
) %>% filter(!match)  %>% nrow()
```

```{r}
# generate random data for code development before data collection
random_data <- as.data.frame(
  lapply(cols_tpsai, function(col) {
    sample(names(choices), size=5, replace=TRUE)
  })
)
colnames(random_data) <- cols_tpsai
```

```{r}
analysis_data <- raw_data # switch to raw_data after data collection

experts_data <- analysis_data %>%
  select(all_of(cols_tpsai))  %>%
  mutate(across(all_of(cols_tpsai), ~ choices[.x]))
```

# Inter-Rater Agreement

```{r}
ratings_matrix <- experts_data %>%
  select(all_of(cols_tpsai)) %>%
  mutate(across(everything(), as.integer)) %>%
  as.matrix() 

overall_alpha <- kripp.alpha(ratings_matrix, method = "nominal")
overall_alpha

interpret_alpha <- case_when(
  overall_alpha$value < 0.20 ~ "Slight agreement",
  overall_alpha$value < 0.40 ~ "Fair agreement",
  overall_alpha$value < 0.60 ~ "Moderate agreement",
  overall_alpha$value < 0.80 ~ "Substantial agreement",
  TRUE                        ~ "Almost perfect agreement"
)

cat("Krippendorff's alpha =", round(overall_alpha$value, 3), "-", interpret_alpha)
```

## Inter-disciplinary expertise

Examining the inter-disciplinary nature of expertise by showing which expert had which expertise.

```{r}
expr_raw <- analysis_data$EXPR

tibble(expert = paste0("E", seq_along(expr_raw)), discipline = expr_raw) %>%
  separate_rows(discipline, sep = ",") %>%
  mutate(discipline = str_trim(discipline)) %>%
  mutate(discipline = fct_infreq(discipline) %>% fct_rev()) %>%  # <- add this
  ggplot(aes(x = expert, y = discipline)) +
  geom_tile(fill = "#378ADD", colour = "white", linewidth = 0.5) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid  = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Expert Evaluations

Expert judges evaluated each item and voted to Keep, Drop, or Reword it. The majority vote determined the final recommendation for each item. When two options tied, the following rules resolved the decision.

1.  Keep and Reword ties favoured Reword on the basis that any doubt about an item warrants revision rather than acceptance.
2.  Drop and Reword ties favoured Drop on the basis that if an item is difficult enough to reword that experts are split, it is better removed.
3.  Keep and Drop ties favoured Keep, giving the item the benefit of the doubt for further evaluation in subsequent validation stages.

```{r}
decisions <- experts_data %>%
  pivot_longer(cols = all_of(cols_tpsai), names_to = "item", values_to = "rating") %>%
  mutate(rating = names(choices)[rating]) %>%
  group_by(item) %>%
  summarise(
    n_keep    = sum(rating == "Keep"),
    n_drop    = sum(rating == "Drop"),
    n_reword  = sum(rating == "Reword"),
    pct_agree = round(max(table(rating)) / n(), 2),
    .groups   = "drop"
  ) %>%
  mutate(
    decision = case_when(
      n_reword > n_keep  & n_reword > n_drop  ~ "Reword",
      n_drop   > n_keep  & n_drop   > n_reword ~ "Drop",
      n_keep   > n_drop  & n_keep   > n_reword ~ "Keep",
      n_keep == n_reword & n_keep > n_drop    ~ "Reword",
      n_drop == n_reword & n_drop > n_keep    ~ "Drop",
      n_keep == n_drop  & n_keep > n_reword   ~ "Keep"
    ),
    item = factor(item, levels = cols_tpsai)
  ) %>%
  arrange(item)

kable(decisions, col.names = c("Item", "Keep", "Drop", "Reword", "% Agreement", "Decision"))
```

## Decisions by theoretical factor

```{r}
factor_summary <- decisions %>%
  mutate(factor = str_extract(item, "^[A-Z]+")) %>%
  group_by(factor) %>%
  summarise(
    keep   = sum(n_keep),
    drop   = sum(n_drop),
    reword = sum(n_reword),
    total  = keep + drop + reword
  ) %>%
  mutate(
    pct_keep   = keep   / total * 100,
    pct_reword = reword / total * 100,
    pct_drop   = drop   / total * 100
  ) %>%
  pivot_longer(
    cols = starts_with("pct_"),
    names_to = "decision",
    names_prefix = "pct_",
    values_to = "percentage"
  ) %>%
  mutate(decision = factor(str_to_title(decision), levels = c("Keep", "Reword", "Drop")))

ggplot(factor_summary, aes(x = factor, y = percentage, fill = decision)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c(
    "Keep"   = "#639922",
    "Reword" = "#BA7517",
    "Drop"   = "#E24B4A"
  )) +
  labs(x = NULL, y = NULL, fill = NULL, title = NULL) +
  scale_y_continuous(labels = scales::label_percent(scale = 1), expand = c(0, 0)) +
  theme_minimal(base_size = 13) +
  theme(legend.position   = "top")
```

# Factors to add

```{r}
analysis_data$COMP_ADD
analysis_data$PERS_ADD
analysis_data$EMOT_ADD
analysis_data$SAFE_ADD
analysis_data$RISK_ADD
analysis_data$CORP_ADD
analysis_data$CRIT_ADD
analysis_data$HUMN_ADD
```
