---
title: "Item Reduction"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

```{r, echo=FALSE}
library(tidyverse)
library(gtools)
library(psych)
library(faux)
library(knitr)

set.seed(8675309) # makes sure random numbers are reproducible
```

```{r}
source("read_qualtrics_data.R")
d <- load_data("data/pilot/TPSai_29+March+2026_15.56.csv", "data/factors.csv")
items_data <- d$items_data
cols_tpsai <- d$item_ids
```

# High missingness

Items where many people respond `Don't know` likely capture constructs that participants are uncertain about and can reduce the effective sample size and item statistics like means and correlations. We decided to drop items exceeding 30% missing responses. Our rationale is that if at least 3 out of 10 individuals were not able to meaningfully engage with an item, dropping it simplifies our scale and allows it to focus on measuring items with generally clearer responses.

Beyond just dropping items with at least 30% missingness, we want to inspect the distribution of missing answers per item. When visually inspecting this distribution we will look for discontinuities, meaning items with a substantially higher proportion of `Don't know` than other items. We may remove such items even if below our pre-set missingness threshold.

```{r}
missing_threshold <- 0.3

missing_data <- items_data %>%
  summarise(
    across(all_of(cols_tpsai), ~ mean(is.na(.x)), .names = "{.col}_missing")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("factor", "q_n"), sep = "_", remove = FALSE)

ggplot(missing_data, aes(x = missing, fill = factor)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = missing_threshold)
```

```{r}
missing_items <- missing_data %>%
  filter(missing > missing_threshold) %>%  # add in other criteria from visual inspection
  select(factor, item, missing) %>% 
  arrange(desc(missing))

if (nrow(missing_items) > 0) {
  missing_items %>% knitr::kable(digits = 2)
}
```

# Distributional Shapes

## Skewness

We examine skewness to check whether any item responses cluster toward one end of the scale.

-   Items with **negative skew** (left tail) may have a "ceiling effect" where an item is too easy to endorse and almost everyone agrees.

-   Items with **positive skew** (right tail) may have a "floor effect" where an item is too extreme to endorse and very few people agree.

Following @kline2016, we inspect for absolute skewness values above 1.5. Items exceeding this threshold will be considered for removal. We also sure the same visual inspect method as high missingness item for the detection of visual discontinuities in skewness values

```{r}
skew_threshold <- 1.5

skew_data <- items_data %>%
  summarise(
    across(all_of(cols_tpsai), ~ psych::skew(.x,    na.rm = TRUE), .names = "{.col}_skew")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("factor", "q_n"), sep = "_", remove = FALSE)

ggplot(skew_data, aes(x = skew, fill = factor)) +
  geom_histogram(binwidth = 0.1) +
  geom_vline(xintercept = skew_threshold) + geom_vline(xintercept = -skew_threshold)
```

```{r}
skew_items <- skew_data %>%
  filter(abs(skew) > skew_threshold) %>%  # add in other criteria from visual inspection
  select(factor, item, skew) %>% 
  arrange(desc(skew))

if (nrow(skew_items) > 0) {
  skew_items %>% knitr::kable(digits = 2)
}
```

## Visual Inspection

Beyond flagging items with high skew, we can also visually inspect the distribution of each item, looking for items where people consistently avoid using certain points in the scale such as items with responses clustered at the top and end of the scale (Bimodal distribution) or items where some scale points are consistently not used. These items might need to be reworded or removed depending on their distributional shape.

```{r}
#| fig-width: 12
#| fig-height: 14
items_data %>% 
  pivot_longer(everything(), names_to = "item", values_to = "response") %>%
  ggplot(aes(x = factor(response))) +
  geom_bar(fill = "#D85A30") +
  scale_x_discrete(limits = as.character(1:7)) +
  facet_wrap(~ item, ncol = 5) +
  labs(x = "", y = "") +
  theme_minimal(base_size = 16)
```
