---
title: "Item Reduction"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

# Data Processing

```{r}
library(tidyverse)
library(gtools)
library(psych)
library(faux)
library(knitr)

set.seed(8675309) # makes sure random numbers are reproducible
```

```{r}
source("read_qualtrics_data.R")
d <- load_data("data/pilot/TPSai_17+March+2026_16.59.csv")
cols_tpsai <- d$cols_tpsai
```

```{r}
# generate random pilot data for code development before data collection
random_data <- as.data.frame(
  lapply(cols_tpsai, function(col) {
    sample(names(d$likert_levels), size=100, replace=TRUE)
  })
)
colnames(random_data) <- cols_tpsai

# switch to d$items_data after data collection
items_data <- random_data %>%
  select(all_of(cols_tpsai))  %>%
  mutate(across(all_of(cols_tpsai), ~ d$likert_levels[.x])) %>% 
  
  # simulate columns with missingness
  faux::messy(prop = .4, "TR_01") %>%
  faux::messy(prop = .2, "DT_06") %>%
  
  # make a high correlation between DISTRUST_01 and _02
  mutate(DT_02 = DT_01 + sample(-1:1, n(), replace = TRUE),
         DT_02 = case_match(DT_02, 0 ~ 1, 6 ~ 5, .default = DT_02))
```

# Single-item Inspection

## High missingness

High missingness items, where many people respond "*Don't know"*, likely capture constructs that participants are uncertain about and can reduce the effective sample size and bias item statistics like means and correlations. Given that we have a substantial number of survey items, removing high missingness items is a way of shortening our survey by focusing on constructs that people have clearer ideas about.

We decided to drop items exceeding 30% missing responses. Our rationale is that if at least 3 out of 10 individuals were not able to meaningfully engage with an item, dropping it simplifies our scale and allows it to focus on measuring items with generally clearer responses. 30% is a relatively moderate threshold for removal of items with high missingness, which might avoid premature dropping of items with moderate missingness may still yield interpretable results.

However, beyond just dropping items with at least 30% missingness, we want to inspect the distribution of missing answers per item, grouped by sub-scale. When visually inspecting this distribution we will look for discontinuities, meaning items with a substantially higher proportion of "Don't know" than other items in the same sub-scale. We may remove such items even if below our pre-set missingness threshold.

```{r}
missing_threshold <- 0.3

missing_data <- items_data %>%
  summarise(
    across(all_of(cols_tpsai), ~ mean(is.na(.x)), .names = "{.col}_missing")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("subscale", "q_n"), sep = "_", remove = FALSE)

ggplot(missing_data, aes(x = missing, fill = subscale)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = missing_threshold)
```

```{r}
missing_items <- missing_data %>%
  filter(missing > missing_threshold) %>%  # add in other criteria from visual inspection
  select(subscale, item, missing)

missing_items %>% knitr::kable(digits = 2)
```

## Distributional Properties

We examine skewness and kurtosis for each item to check whether responses are spread across the scale or piled up at one end. Items where most people gave the same answer have little variation, which makes it harder for factor analysis to detect the underlying structure, regardless of the estimation method used.

Following @kline2016, we treat absolute skewness and kurtosis values above 1.5 as a sign that an item's responses are too bunched together to be useful. Items exceeding this threshold will be considered for removal alongside the other item reduction criteria. Additionally, we may consider keeping items with high skew or kurtosis for comparison or reporting purposes but flag them for removal from factor analysis.

```{r}
dist_threshold <- 1.5

dist_data <- items_data %>%
  summarise(
    across(all_of(everything()), list(
        skew = ~ psych::skew(.x,    na.rm = TRUE),
        kurt = ~ psych::kurtosi(.x, na.rm = TRUE)
      ), .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("subscale", "q_n"), sep = "_", remove = FALSE)

ggplot(dist_data, aes(x = skew, y = kurt, colour = subscale, label = item)) +
  geom_point() +
  geom_hline(yintercept = c(-dist_threshold, dist_threshold), linetype = "dashed") +
  geom_vline(xintercept = c(-dist_threshold, dist_threshold), linetype = "dashed") +
  labs(x = "Skewness", y = "Kurtosis")
```

```{r}
dist_items <- dist_data %>%
  filter(abs(skew) > dist_threshold | abs(kurt) > dist_threshold) %>% 
  select(subscale, item, skew, kurt)

dist_items  %>% knitr::kable(digits = 2)
```

Let's also inspect how the distributions of flagged items actually look like

```{r}
flagged_items <- dist_items %>% pull(item)

items_data %>% 
  select(all_of(flagged_items)) %>%
  pivot_longer(everything(), names_to = "item", values_to = "response") %>%
  left_join(dist_items, by = "item") %>%
  ggplot(aes(x = factor(response))) +
  geom_bar(fill = "#D85A30") +
  facet_wrap(~ item) +
  labs(x = "Response", y = "Count") +
  theme_minimal()
```

## Item-Total Correlations

For each item, we compute the *adjusted item-total correlation*, the correlation between that item and the sum of all other items in the same sub-scale, excluding itself. Items with an adjusted item-total correlation below .30 are flagged for potential removal, as they contribute little to what the sub-scale is measuring overall [@boateng2018]

As with previous checks, we will also visually inspect the distribution of item-total correlations within each sub-scale, looking for discontinuities that may identify items worth removing even if they fall above our pre-set threshold. Particular attention will be paid to any negatively correlating items, which would suggest an item is moving in the opposite direction to the rest of its sub-scale — a strong signal of potential issues with wording or construct alignment.

```{r}
it_cor_threshold <- 0.3

it_cor_data <- list(
  TR    = d$cols_trust,
  DT = d$cols_distrust,
  CD  = d$cols_company
) %>%
  map_dfr(~ {
    items_data %>%
      select(all_of(.x)) %>%
      alpha() %>%
      .$item.stats %>%
      rownames_to_column("item") %>%
      select(item, it_corr = r.drop)
  }, .id = "subscale")

it_cor_data %>%
  ggplot(aes(x = it_corr, fill= subscale)) +
  facet_wrap(~subscale, ncol = 1, scales = "free_y") +
  geom_histogram(binwidth = 0.05) +
  geom_vline(xintercept = it_cor_threshold) +
  geom_vline(xintercept = -it_cor_threshold) +
  scale_x_continuous(breaks = seq(-1, 1, .2)) +
  coord_cartesian(xlim = c(-1, 1))
```

```{r}
it_cor_items <- it_cor_data %>%
  filter(abs(it_corr) < it_cor_threshold) %>%
  select(subscale, item, it_corr)

it_cor_items %>%
  knitr::kable(digits = 2)
```

## Summary of single items to iinspect

```{r}
options(knitr.kable.NA = " ")

missing_items %>%
  full_join(dist_items, by=c("item", "subscale")) %>% 
  full_join(it_cor_items, by=c("item", "subscale")) %>% 
  select(subscale, item, missing, skew, kurt, it_corr) %>% 
  arrange(item) %>% 
  knitr::kable(digits = 2)
```


# Item pairs: Inter-item Correlations

Next, we examine pairwise Polychoric inter-item correlations to identify redundant items. Polychoric correlations are used as they are more appropriate for ordinal Likert-scale data than Pearson correlations @boateng2018. Items correlating above .80 with another item are considered redundant, as this indicates the two items share more than 64% of their variance (*r*² = .64), leaving little unique contribution to the scale.

As per our missingness check, we will also visually inspect inter-item correlations for any discontinuity in distribution across sub-scales, potentially identifying items worth merging even if correlated under our pre-set threshold.

We examine inter-item correlations within and across sub-scales.

```{r}
ii_cor_threshold <- 0.8

iia_cor_data <- 
  items_data %>%
  select(all_of(cols_tpsai)) %>%
  polychoric() %>%
  .$rho %>%
  as.data.frame() %>%
  rownames_to_column("item1") %>%
  pivot_longer(-item1, names_to = "item2", values_to = "r") %>%
  filter(r > -1 & r < 1) %>%
  filter(item1 < item2)

iia_cor_data %>%
  ggplot(aes(x = r)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = ii_cor_threshold) +
  geom_vline(xintercept = -ii_cor_threshold) +
  scale_x_continuous(breaks = seq(-1, 1, .2)) +
  coord_cartesian(xlim = c(-1, 1))
```

```{r}
iia_cor_data %>%
  filter(abs(r) > ii_cor_threshold) %>%
  select(item1, item2, r) %>% # add in other criteria from visual inspection
  arrange(r) %>%
  knitr::kable(digits = 2)
```

