---
title: "Item Reduction"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

```{r, echo=FALSE}
library(tidyverse)
library(gtools)
library(psych)
library(faux)
library(knitr)

set.seed(8675309) # makes sure random numbers are reproducible
```

```{r}
source("read_qualtrics_data.R")
d <- load_data("data/pilot/TPSai_29+March+2026_15.56.csv", "data/factors.csv")
items_data <- d$items_data
cols_tpsai <- d$item_ids
```

# High missingness

Items where many people respond `Don't know` likely capture constructs that participants are uncertain about and can reduce the effective sample size and item statistics like means and correlations. We decided to drop items exceeding 30% missing responses. Our rationale is that if at least 3 out of 10 individuals were not able to meaningfully engage with an item, dropping it simplifies our scale and allows it to focus on measuring items with generally clearer responses.

Beyond just dropping items with at least 30% missingness, we want to inspect the distribution of missing answers per item. When visually inspecting this distribution we will look for discontinuities, meaning items with a substantially higher proportion of `Don't know` than other items. We may remove such items even if below our pre-set missingness threshold.

```{r}
missing_threshold <- 0.3

missing_data <- items_data %>%
  summarise(
    across(all_of(cols_tpsai), ~ mean(is.na(.x)), .names = "{.col}_missing")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("factor", "q_n"), sep = "_", remove = FALSE)

ggplot(missing_data, aes(x = missing, fill = factor)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = missing_threshold)
```

```{r}
missing_items <- missing_data %>%
  filter(missing > missing_threshold) %>%  # add in other criteria from visual inspection
  select(factor, item, missing) %>% 
  arrange(desc(missing))

if (nrow(missing_items) > 0) {
  missing_items %>% knitr::kable(digits = 2)
}
```

# Distributional Shapes

## Visual Inspection

Let us start by visually inspecting the distribution of each item. Here we are looking for items where people consistently avoid using certain points in the scale such as items with responses clustered at the top and end of the scale (Bimodal distribution) or items where some scale points are consistently not used. These items might need to be reworded or removed depending on their distributional shape.

Moreover, we can look for items that have a remarkably visually different distribution that all other items in the same factor, suggesting that the item might not belong to the intended factor and we might need to remove it or reword it.

```{r}
#| fig-width: 12
#| fig-height: 14
items_data %>% 
  pivot_longer(everything(), names_to = "item", values_to = "response") %>%
  ggplot(aes(x = factor(response))) +
  geom_bar(fill = "#D85A30") +
  scale_x_discrete(limits = as.character(1:7)) +
  facet_wrap(~ item, ncol = 5) +
  labs(x = "", y = "") +
  theme_minimal(base_size = 16)
```

## Skewness

We examine skewness to check whether any item responses cluster toward one end of the scale.

-   Items with **negative skew** (left tail) may have a "ceiling effect" where an item is too easy to endorse and almost everyone agrees.

-   Items with **positive skew** (right tail) may have a "floor effect" where an item is too extreme to endorse and very few people agree.

Following @kline2016, we inspect for absolute skewness values above 1.5. Items exceeding this threshold will be considered for removal. We also sure the same visual inspect method as high missingness item for the detection of visual discontinuities in skewness values

```{r}
skew_threshold <- 1.5

skew_data <- items_data %>%
  summarise(
    across(all_of(cols_tpsai), ~ psych::skew(.x,    na.rm = TRUE), .names = "{.col}_skew")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("factor", "q_n"), sep = "_", remove = FALSE)

ggplot(skew_data, aes(x = skew, fill = factor)) +
  geom_histogram(binwidth = 0.1) +
  geom_vline(xintercept = skew_threshold) + geom_vline(xintercept = -skew_threshold)
```

```{r}
skew_items <- skew_data %>%
  filter(abs(skew) > skew_threshold) %>%  # add in other criteria from visual inspection
  select(factor, item, skew) %>% 
  arrange(desc(skew))

if (nrow(skew_items) > 0) {
  skew_items %>% knitr::kable(digits = 2)
}
```

We can visually explore these flagged items in the above plots of distributions.

# Inter-item Correlations

Next, we examine pairwise inter-item correlations to identify redundant items. Items correlating above .80 with another item are considered redundant, as this indicates the two items share more than 64% of their variance (*r*² = .64), leaving little unique contribution to the scale. As per our missingness check, we will also visually inspect inter-item correlations for any discontinuity in distribution across sub-scales, potentially identifying items worth merging even if correlated under our pre-set threshold.

We intend to use Polychoric correlations as they are more appropriate for ordinal Likert-scale data than Pearson correlations [@boateng2018].

```{r}
ii_cor_threshold <- 0.8

iia_cor_data <- 
  items_data %>%
  select(all_of(cols_tpsai)) %>%
  #TODO: using Pearson just for tester data, switch to `polychoric() %>%.$rho` after
  cor(use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("item1") %>%
  pivot_longer(-item1, names_to = "item2", values_to = "r") %>%
  filter(r > -1 & r < 1) %>%
  filter(item1 < item2)

iia_cor_data %>%
  ggplot(aes(x = r)) +
  geom_histogram(binwidth = 0.01, fill = "steelblue") +
  geom_vline(xintercept = ii_cor_threshold) + geom_vline(xintercept = -ii_cor_threshold) +
  scale_x_continuous(breaks = seq(-1, 1, .2)) +
  coord_cartesian(xlim = c(-1, 1))
```

```{r}
item_factor_lookup <- tibble(
  item = cols_tpsai,
  factor = str_extract(item, "^[A-Z]+")
)

iia_cor_data <- iia_cor_data %>%
  left_join(item_factor_lookup, by = c("item1" = "item")) %>%
  rename(factor1 = factor) %>%
  left_join(item_factor_lookup, by = c("item2" = "item")) %>%
  rename(factor2 = factor) %>%
  mutate(pair_type = if_else(factor1 == factor2, "Same factor", "Cross factor")) %>%
  select(item1, item2, r, pair_type)
```

## Positive within-factor

Items are moving together as expected. However, if r exceeds our redundancy threshold, the items may be asking the same question twice. We might dropping the weaker item (lower item-total correlation) or merging the pair into a single composite.

```{r}
iia_cor_data %>%
  filter(r > ii_cor_threshold) %>%
  filter(pair_type == "Same factor") %>%
  select(-pair_type) %>%
  knitr::kable(digits = 2)
```

## Negative within-factor

One or more items are worded in the opposite direction to the rest of the subscale. Check for items that need reverse-scoring and recode before re-running.

```{r}
iia_cor_data %>%
  filter(r < -ii_cor_threshold) %>%
  filter(pair_type == "Same factor") %>%
  select(-pair_type) %>%
  knitr::kable(digits = 2)
```

## Across-factors

Two subscales are overlapping more than expected. Consider whether the items belong to the same factor, or whether the two subscales need clearer conceptual separation. If persistent across multiple item pairs, review the factor structure.

```{r}
iia_cor_data %>%
  filter(abs(r) > ii_cor_threshold) %>%
  filter(pair_type == "Cross factor") %>%
  select(-pair_type) %>%
  knitr::kable(digits = 2)
```
