---
title: "Item Reduction"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

```{r, echo=FALSE}
library(tidyverse)
library(gtools)
library(psych)
library(faux)
library(knitr)

set.seed(8675309) # makes sure random numbers are reproducible
```

```{r}
source("scripts/read_qualtrics_data.R")
d <- load_data("data/TPSai+-+Pilot+Tester_27+April+2026_13.02.csv", "data/factors_v1.1.csv")
items_data <- d$items_data
cols_tpsai <- d$item_ids
nrow(items_data)
```

# High missingness

Items where many people respond `Don't know` likely capture constructs that participants are uncertain about and can reduce the effective sample size and item statistics like means and correlations. We decided to drop items exceeding 30% missing responses. Our rationale is that if at least 3 out of 10 individuals were not able to meaningfully engage with an item, dropping it simplifies our scale and allows it to focus on measuring items with generally clearer responses.

Beyond just dropping items with at least 30% missingness, we want to inspect the distribution of missing answers per item. When visually inspecting this distribution we will look for discontinuities, meaning items with a substantially higher proportion of `Don't know` than other items. We may remove such items even if below our pre-set missingness threshold.

```{r}
missing_threshold <- 0.3

missing_data <- items_data %>%
  summarise(
    across(all_of(cols_tpsai), ~ mean(is.na(.x)), .names = "{.col}_missing")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("factor", "q_n"), sep = "_", remove = FALSE)

# adding another threshold at 0.1 from visual inspection
v_missing_threshold <- 0.1

ggplot(missing_data, aes(x = missing, fill = factor)) +
  geom_histogram(binwidth = 0.0075) +
  geom_vline(xintercept = missing_threshold) +
  geom_vline(xintercept = v_missing_threshold, linetype="dashed",color='red')
```

5% threshold means more than `{r} nrow(items_data) * 0.05` missing responses out of `{r} nrow(items_data)`.

```{r}
missing_items <- missing_data %>%
  filter(missing > missing_threshold | missing > v_missing_threshold) %>%
  select(item, missing) %>% 
  arrange(desc(missing))

if (nrow(missing_items) > 0) {
  missing_items %>% knitr::kable(digits = 2)
}
```

# Distributional Shapes

## Skewness and Kurtosis

We examine skewness and kurtosis for each item to check whether responses are spread across the scale or piled up at one end. If everyone scores high/low on an item, it means that the item is too easy/hard to endorse and cannot discriminate between people with high versus low levels of that trait being measured. Moreover, items where most people gave the same answer have little variation, which makes it harder for factor analysis to detect the underlying structure, regardless of the estimation method used.

-   Items with **negative skew** and **high positive kurtosis** (left tail) may have a "ceiling effect" where an item is too easy to endorse and almost everyone agrees.

-   Items with **positive skew** and **high negative kurtosis**(right tail) may have a "floor effect" where an item is too extreme to endorse and very few people agree.

Avoiding ceiling/floor effects is crucial because if items cluster at the extremes, you lose the ability to distinguish among people at those ends.

Following @kline2016, we treat absolute skewness and kurtosis values above 1.5 as a sign that an item's responses are too bunched together to be useful. Items exceeding this threshold will be considered for removal or rewording alongside the other item reduction criteria. Additionally, we may consider keeping items with high skew or kurtosis for comparison or reporting purposes but flag them for removal from factor analysis.

```{r}
dist_threshold <- 1.5

dist_data <- items_data %>%
  summarise(
    across(all_of(everything()), list(
        skew = ~ psych::skew(.x,    na.rm = TRUE),
        kurt = ~ psych::kurtosi(.x, na.rm = TRUE)
      ), .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("subscale", "q_n"), sep = "_", remove = FALSE)

# adding another threshold at 1.25 from visual inspection
v_dist_threshold <- 1.25

ggplot(dist_data, aes(x = skew, y = kurt, colour = subscale, label = item)) +
  geom_point() +
  geom_hline(yintercept=c(-dist_threshold, dist_threshold)) +
  geom_vline(xintercept=c(-dist_threshold, dist_threshold)) +
  geom_hline(yintercept=c(-v_dist_threshold, v_dist_threshold),linetype="dashed",color='red') +
  geom_vline(xintercept=c(-v_dist_threshold, v_dist_threshold),linetype="dashed",color='red') +
  labs(x = "Skewness", y = "Kurtosis")
```

```{r}
dist_items <- dist_data %>%
  filter(abs(skew) > v_dist_threshold | abs(kurt) > v_dist_threshold) %>% 
  select(item, skew, kurt)

if (nrow(dist_items) > 0) {
  dist_items  %>% knitr::kable(digits = 2)
}
```

## Visual Inspection

Beyond flagging items with high skew, we can also visually inspect the distribution of each item, looking for items where people consistently avoid using certain points in the scale such as items with responses clustered at the top and end of the scale (Bimodal distribution) or items where some scale points are consistently not used. These items might need to be reworded or removed depending on their distributional shape.

```{r, warning=FALSE}
#| fig-height: 2.5
#| fig-width: 10
long_data <- items_data %>%
  pivot_longer(everything(), names_to = "item", values_to = "response") %>%
  mutate(factor = sub("_.*", "", item))

y_max <- long_data %>% 
  count(item, response) %>% 
  group_by(item) %>% 
  mutate(prop = n / sum(n)) %>% 
  pull(prop) %>% 
  max()

for (f in unique(long_data$factor)) {
  p <- long_data %>%
    filter(factor == f) %>%
    ggplot(aes(x = factor(response), y = after_stat(prop), group = 1)) +
    geom_bar(fill = "#D85A30") +
    scale_x_discrete(limits = as.character(1:7)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    coord_cartesian(ylim = c(0, y_max)) +
    facet_wrap(~ item, nrow = 1) +
    labs(x = "", y = "") +
    theme_minimal(base_size = 16)
  
  print(p)
}
```

# Item Statistics

To complement the diagnostic checks above, we compile a single table of descriptive statistics for all items. This provides an at-a-glance reference showing sample size, central tendency, dispersion, and distributional shape for each item, alongside the missingness and skewness values used in our screening decisions.

```{r}
item_stats <- items_data %>%
  select(all_of(cols_tpsai)) %>%
  pivot_longer(everything(), names_to = "item", values_to = "response") %>%
  group_by(item) %>%
  summarise(
    n        = sum(!is.na(response)),
    mean     = mean(response, na.rm = TRUE),
    sd       = sd(response, na.rm = TRUE),
    median   = median(response, na.rm = TRUE),
    min      = suppressWarnings(min(response, na.rm = TRUE)),
    max      = suppressWarnings(max(response, na.rm = TRUE)),
    skew     = psych::skew(response, na.rm = TRUE),
    kurtosis = psych::kurtosi(response, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  arrange(item)

item_stats %>%
  knitr::kable(
    digits = 2,
    col.names = c("Item", "N", "Mean", "SD",
                  "Median", "Min", "Max", "Skew", "Kurtosis"),
  )
```

# Survey Duration

```{r, warning=FALSE}
ggplot(d$raw_data, aes(x = `Duration (in seconds)`)) +
  geom_histogram(bins = 50) +
  xlim(NA, quantile(d$raw_data$`Duration (in seconds)`, 0.95, na.rm = TRUE)) +
  geom_vline(xintercept = 600)
```
