---
title: "Exploratory Factor Analysis (EFA)"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

# Data Preparation

```{r, echo=FALSE}
library(tidyverse)
library(gtools)
library(psych)
library(knitr)
library(parameters)
```

```{r}
options(readr.show_col_types = FALSE)
file_path <- "data/TPSai+-+Pilot.csv" #TODO: replace pilot data with EFA sample
raw_data  <- read_csv(file_path, n_max = 2)
colnames  <- names(raw_data)
q_text    <- setNames(sub(".* - ", "", as.character(raw_data[1, ])), colnames)
raw_data  <- read_csv(file_path, col_names = colnames, skip = 3)
```

```{r}
factors_map <- read_csv('surveys/pilot-test_qs.csv') #TODO: replace with EFA questions
cols_tpsai <- sprintf("TPSAI_%d", seq_along(seq_along(1:nrow(factors_map))))
```

```{r}
# check that survey and factors correspond
data.frame(
  survey_col  = cols_tpsai,
  survey_q    = unname(q_text[cols_tpsai]),
  factor_item = factors_map$Item
) %>%
  mutate(match = survey_q == factor_item) |>
  filter(!match)  %>% nrow() == 0
```

```{r}
likert_levels <- c(
  "Strongly disagree"          = 1,
  "Disagree"                   = 2,
  "Somewhat disagree"          = 3,
  "Neither agree nor disagree" = 4,
  "Somewhat agree"             = 5,
  "Agree"                      = 6,
  "Strongly agree"             = 7,
  "Don't know"                 = NA
)

efa_data <- raw_data |>
  mutate(across(all_of(cols_tpsai), ~ likert_levels[.x])) |>
  select(all_of(cols_tpsai))
```

```{r}
# Rename columns TPSAI_1 through TPSAI_9 to TPSAI_01 through TPSAI_09
names(efa_data) <- gsub("^TPSAI_(\\d)$", "TPSAI_0\\1", names(efa_data))
cols_tpsai <- gsub("^TPSAI_(\\d)$", "TPSAI_0\\1", cols_tpsai)
```

```{r}
# up-sample pilot data to N=300 with simulation
# use only if need a bigger N to develop analysis
# drop for EFA data
source("scripts/simulate_data.R")
efa_data <- simulate_data(efa_data, n=300)
```

**Exploratory Factor Analysis** (EFA) identifies hidden structures in a set of items. The core idea: if items correlate, they're probably measuring a common underlying thing — a latent factor. EFA finds the factors that can reproduce the observed correlations, retaining the solution that leaves the smallest residuals. Throughout our EFA, we use polychoric correlation matrix rather than the standard Pearson correlation matrix, as polychoric correlations are more appropriate for ordinal Likert data [@boateng2018].

```{r}
R <- polychoric(efa_data %>% select(all_of(cols_tpsai)))$rho
n_obs <- nrow(efa_data)
```

**Helper functions**

```{r}
plot_loadings <- function(loadings) {
  loads <- loadings |>
    unclass() |>
    as.data.frame() |>
    rownames_to_column("item") |>
    pivot_longer(
      cols = -c(item), 
      names_to = "factor", 
      values_to = "loading") |>
    mutate(
      item = factor(item),
      label = as.character(round(loading, 2))
    ) |>
    group_by(item) |>
    mutate(bold = abs(loading) == max(abs(loading))) |>
    ungroup()
  
  ggplot(loads, aes(x = factor, y = item, fill = loading)) +
    geom_tile() +
    geom_text(data = filter(loads, !bold),
              aes(label = label, colour = abs(loading) > 0.5)) +
    geom_text(data = filter(loads, bold),
              aes(label = label, colour = abs(loading) > 0.5),
              fontface = "bold") +
    scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "black"), guide = "none") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
    scale_y_discrete(limits = rev) +
    theme_minimal() +
    labs(x = "", y = "", fill = "")
}
```

```{r}
factors_as_items <- function(efa) {
  cfa_syntax <- paste(capture.output(efa_to_cfa(efa)), collapse = "\n")
  lines <- strsplit(cfa_syntax, "\n")[[1]]
  
  rows <- list()
  for (line in lines) {
    factor_name <- regmatches(line, regexpr("MR\\d+", line))
    if (length(factor_name) == 0) next
    
    nums <- as.integer(regmatches(line, gregexpr("(?<=TPSAI_)\\d+", line, perl = TRUE))[[1]])
    
    rows[[length(rows) + 1]] <- data.frame(
      Factor = factor_name,
      IID    = nums,
      Item   = factors_map[nums, "Item"],
      stringsAsFactors = FALSE
    )
  }
  
  do.call(rbind, rows) |>
    arrange(Factor, IID)
}
```

# Top-Level Structure

Our theoretical assumption is that TPSai items are primarily organised into a top-level structure with the two factors of *Trust* and *Distrust.* Thus, the first step in our EFA is it explore the data we collected in its entirely, confirming whether the top-level Trust/Distrust factor structure exists or whether an entirely different emergent structure exists at the top level.

## Factorability

First, we report the Kaiser-Meyer-Olkin (KMO) index which asks whether the correlations among items are sufficiently compact and patterned to yield reliable factors. We compute both the overall measure of sampling adequacy (MSA) and the minimum per-item MSA.

```{r}
kmo <- KMO(R)
kmo$MSA
```

If the overall MSA is below 0.8, we inspect per-item MSA and drop any items with MSA \< 0.5, then recompute the overall MSA. If the overall MSA remains below 0.8 after dropping problematic items, we will collect additional data before proceeding.

```{r}
tibble(
  item = names(kmo$MSAi),
  msa  = round(kmo$MSAi, 3)
) |>
  arrange(msa) |>
  kable(col.names = c("Item", "MSA"))
```

## How Many Factors?

We use multiple criteria to estimate the number of factors at the top level, as no single method is universally reliable. We report parallel analysis, inspect the scree plot, the Velicer MAP test, and VSS complexity indices.

**Parallel analysis** compares each factor's eigenvalue from the real data against the mean eigenvalue from randomly generated correlation matrices of the same dimensions. Eigenvalues measure how much shared variance a factor accounts for. A factor is retained if its eigenvalue exceeds the random baseline, meaning it captures more variance than would be expected by chance alone.

We also inspect the **scree plot** generated by parallel analysis for a clear elbow, the point at which the drop in eigenvalues between consecutive factors becomes substantially smaller, indicating that remaining factors contribute little additional explanatory power.

```{r}
fa.parallel(
  R,
  n.obs = nrow(efa_data),
  fm = "minres", # minimum residual: no distributional assumptions, suits ordinal data
  fa = "fa",     # factor analysis (not PCA)
)
```

The **Velicer MAP test** iteratively partials out each factor from the correlation matrix and computes the average squared partial correlation. The optimal number of factors is the point at which this statistic is minimised, reflecting the solution where common variance is most fully extracted and only unique variance remains.

**Very Simple Structure (VSS)** evaluates how well a simplified factor solution, where each item loads on only one factor, reproduces the original correlation matrix. We report VSS complexity 1 and complexity 2 for completeness, but do not use them as primary decision criteria.

```{r}
vss(R, n.obs = nrow(efa_data), rotate = "oblimin", fm = "minres")
```

**Decision logic**

We use the MAP test and scree plot elbow as our primary criteria. If both agree, we adopt that solution. If they disagree, we examine parallel analysis and VSS complexity 2 as additional evidence. If no clear convergence emerges across all criteria, we adopt the solution suggested by MAP, which is generally considered the most accurate criterion for psychological scale data [@velicer1976; @zwick1986].

```{r}
# Set number of factors based on MAP and scree plot
n_factors <- 2  # replace with chosen number after inspecting criteria
```

Based on this, we arrive at a suggested number of factors and proceed as follows:

- **n = 1**: suggests a unidimensional structure without clear separation between Trust and Distrust. We do not proceed with the Trust/Distrust model.

- **n = 2**: suggests overlap with the Trust/Distrust model. We proceed to the EFA solution.

- **n \> 2**: suggests an entirely new factor structure. We proceed to the EFA solution and investigate the emergent structure.

## EFA Solution

We extract the suggested number of factors using **MINRES** (minimum residual) estimation with **oblimin** rotation. MINRES makes no distributional assumptions about the data, making it appropriate for ordinal Likert items. Oblimin is an oblique rotation method that allows factors to correlate — a realistic assumption for psychological constructs such as Trust and Distrust, which are unlikely to be fully independent. Oblimin also simplifies the loading structure, making factors easier to interpret.

```{r}
efa <- fa(
  r = R,
  nfactors = n_factors,  # determined in previous section
  n.obs = nrow(efa_data),
  fm = "minres",          # minimum residual: no distributional assumptions
  rotate = "oblimin"      # oblique rotation: allows factors to correlate
)
```

We visualise the factor loadings as a heatmap, where each cell represents the strength and direction of the relationship between an item and a factor. Loadings close to 1 or -1 indicate a strong relationship; loadings close to 0 indicate little relationship. Bold values indicate the highest loading for that item across all factors.

```{r}
#| fig-width: 7
#| fig-height: 8
plot_loadings(efa$loadings)
```

To interpret the factor structure, we examine which items load most strongly on each factor. Where the structure is emergent, we use item content to identify the underlying construct each factor represents.

```{r}
factors_as_items(efa) |> kable()
```

Where the factor structure maps onto a theoretical model, either the Trust/Distrust (n = 2) or the 4F/8F models (n \> 2), we additionally check whether items load on the correct theoretical factor. Items loading primarily on the wrong factor are candidates for removal or reversal:

- **Removal**: if the item's content is not theoretically consistent with the factor it loads on
- **Reversal**: if the item's content is theoretically consistent with the factor it loads on, but the loading is negative

*The following example uses pilot data to illustrate how these criteria are applied in practice. The actual decisions will be made on the EFA sample.*

In the pilot data, items 24, 27, 31, and 34 load primarily on the opposite factor to their theoretical assignment and are candidates for removal.

### Item Reduction

We inspect the loading pattern for simple structure — ideally each item loads strongly on one factor and weakly on all others [@buchanan2014]. We flag items for removal based on two empirical criteria:

- **Low primary loading**: primary loading \< 0.4 — the item does not relate strongly enough to any factor
- **Cross-loading**: gap between primary and secondary loading \< 0.05 — the item does not discriminate sufficiently between factors. We use a lenient threshold of 0.05 rather than the conventional 0.10, acknowledging that Trust and Distrust are genuinely correlated constructs and some cross-loading is expected.

Both criteria are applied independently — an item is flagged if it fails either criterion. Items near the thresholds are subject to theoretical review and may be retained if their content is theoretically justified.

```{r}
min_primary <- 0.4
min_diff    <- 0.05

loadings_df <- efa$loadings |>
  unclass() |>
  as.data.frame() |>
  rownames_to_column("item") |>
  rowwise() |>
  mutate(
    abs_loads = list(sort(abs(c_across(-item)), decreasing = TRUE)),
    primary   = abs_loads[[1]],
    secondary = abs_loads[[2]],
    diff      = primary - secondary,
    flagged   = primary < min_primary | diff < min_diff
  ) |>
  ungroup()

ggplot(loadings_df, aes(primary, diff, colour = flagged)) +
  geom_vline(xintercept = min_primary, linetype = "dashed", colour = "grey60") +
  geom_hline(yintercept = min_diff,   linetype = "dashed", colour = "grey60") +
  geom_point(size = 2) +
  geom_text(aes(label = str_remove(item, "TPSAI_")),
            size = 3, hjust = -0.4, vjust = 0.5, show.legend = FALSE) +
  scale_colour_manual(values = c(`FALSE` = "#1D9E75", `TRUE` = "#D85A30"),
                      labels = c("Passed", "Flagged")) +
  labs(x = "Primary loading", y = "Primary − secondary gap", colour = NULL) +
  theme_minimal()
```

Items falling below the vertical dashed line have a low primary loading; items falling below the horizontal dashed line are cross-loaders. Items in the lower-left quadrant fail both criteria.

```{r}
loadings_df |>
  filter(flagged) |>
  select(item, primary, diff) |>
  kable(
    digits     = 2,
    col.names  = c("Item", "Primary", "Gap")
  )
```

Items not flagged by any criterion are retained. Items flagged by one or more criteria are candidates for removal, subject to theoretical review. After removing flagged items, we update the item list and proceed to the substructure analysis.

*In the pilot data, the following items may be dropped:*

```{r}
items_to_drop <- c(
  "TPSAI_24", "TPSAI_27", "TPSAI_31", "TPSAI_34", # wrong factor
  "TPSAI_23", "TPSAI_32" # low or cross loaders
)
cols_tpsai <- setdiff(cols_tpsai, items_to_drop)
```

# Substructures

We proceed to substructure analysis only if the top-level EFA confirms a two-factor Trust/Distrust structure. We examine whether meaningful substructure exists within each factor using a **bifactor model**, which simultaneously estimates a general factor (shared by all items within the block) and specific group factors (capturing variance unique to subgroups of items).

We apply this separately to Trust and Distrust items, using the theoretical item assignments to define each block. After item reduction, `cols_tpsai` contains only surviving items, so the blocks are defined as:

```{r}
trust_items    <- cols_tpsai[cols_tpsai %in% sprintf("TPSAI_%02d", 1:19)]
distrust_items <- cols_tpsai[cols_tpsai %in% sprintf("TPSAI_%02d", 20:39)]

R_trust    <- R[trust_items, trust_items]
R_distrust <- R[distrust_items, distrust_items]
```

We fit bifactor models with k = 2, 3, and 4 group factors for each block, covering the range between our 4F theoretical model (2 subgroups per factor) and our 8F theoretical model (4 subgroups per factor). The maximum k tested is additionally constrained by block size. We do not test values of k where floor(n_items / 3) \< k, as this would make the minimum items per factor criterion unachievable by design.

A solution is acceptable only if **all** minimum requirements are met:

- **RMSEA \< 0.10** — acceptable model fit
- **Minimum 3 items per group factor** — each group factor must be sufficiently represented
- **Max/min items ratio ≤ 3** — group factors must be reasonably balanced in size
- **ω_h \< 0.90** — the general factor must not dominate to the point where group factors are trivial
- **≤ 20% of items are cross-loaders** — where a cross-loader is defined as an item whose gap between primary and secondary group factor loading is \< 0.05

Where multiple solutions meet all minimum requirements, we prefer the solution with the fewest cross-loaders. Where cross-loaders are equal, we prefer the solution with the best factor balance (lowest ratio). Where both are equal, we prefer the solution with the lower BIC. If criteria point in different directions and no clear winner emerges, we report both solutions and discuss the trade-offs.

We compare the selected solution qualitatively against our theoretical models — Cognitive/Affective and Artefact/Corporate for the 2F model, or the eight specific dimensions for the 8F model. The substructure may also be emergent, in which case we interpret it based on item content and report it as exploratory.

```{r}
compare_omega <- function(omega_list, labels = c("k=2", "k=3", "k=4"),
                          gap_threshold = 0.05, min_loading = 0.2) {
  
  map_dfr(seq_along(omega_list), function(i) {
    om  <- omega_list[[i]]
    sl  <- om$schmid$sl
    
    # Extract group factor columns only (exclude g, h2, u2, p2)
    group_cols <- sl[, grepl("^F", colnames(sl)), drop = FALSE]
    
    # Assign items to primary factor only if loading >= min_loading
    primary_factor <- apply(group_cols, 1, function(x) {
      if (max(abs(x)) < min_loading) return(NA)
      which.max(abs(x))
    })
    
    items_per_factor <- table(primary_factor[!is.na(primary_factor)])
    
    # Cross-loaders: gap between primary and secondary group factor loading < gap_threshold
    cross_loaders <- sum(apply(group_cols, 1, function(x) {
      sorted <- sort(abs(x), decreasing = TRUE)
      if (length(sorted) < 2) return(FALSE)
      (sorted[1] - sorted[2]) < gap_threshold
    }))
    
    n_items <- nrow(group_cols)
    
    tibble(
      model         = labels[[i]],
      RMSEA         = round(om$schmid$RMSEA["RMSEA"], 3),
      BIC           = round(om$schmid$BIC, 1),
      omega_h       = round(om$omega_h, 3),
      min_items     = min(items_per_factor),
      max_items     = max(items_per_factor),
      ratio         = round(max(items_per_factor) / min(items_per_factor), 1),
      cross_loaders = cross_loaders,
      cross_pct     = round(cross_loaders / n_items * 100, 1)
    )
  })
}
```

## Trust Substructure

```{r}
omega_trust_2 <- omega(
  m        = R_trust,
  nfactors = 2,
  n.obs    = nrow(efa_data),
  fm       = "minres",
  rotate   = "oblimin"
)

omega_trust_3 <- omega(
  m        = R_trust,
  nfactors = 3,
  n.obs    = nrow(efa_data),
  fm       = "minres",
  rotate   = "oblimin"
)

omega_trust_4 <- omega(
  m        = R_trust,
  nfactors = 4,
  n.obs    = nrow(efa_data),
  fm       = "minres",
  rotate   = "oblimin"
)
```

```{r}
compare_omega(list(omega_trust_2, omega_trust_3, omega_trust_4)) |> kable()
```

## Distrust Substructure

```{r}
omega_distrust_2 <- omega(
  m        = R_distrust,
  nfactors = 2,
  n.obs    = nrow(efa_data),
  fm       = "minres",
  rotate   = "oblimin"
)

omega_distrust_3 <- omega(
  m        = R_distrust,
  nfactors = 3,
  n.obs    = nrow(efa_data),
  fm       = "minres",
  rotate   = "oblimin"
)

omega_distrust_4 <- omega(
  m        = R_distrust,
  nfactors = 4,
  n.obs    = nrow(efa_data),
  fm       = "minres",
  rotate   = "oblimin"
)
```

```{r}
compare_omega(list(omega_distrust_2, omega_distrust_3, omega_distrust_4)) |> kable()
```
