---
title: "Factor Analysis"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

```{r, echo=FALSE}
library(tidyverse)
library(gtools)
library(psych)
library(knitr)
library(parameters)

seed <- 8675309
set.seed(seed) # makes sure random numbers are reproducible
```

```{r}
source("simulate_data.R")

efa_data <- simulate_data(
  sample_path = "data/pilot/TPSai_29+March+2026_15.56.csv", 
  factors_path =  "data/factors.csv",
  seed = seed,
  n = 250)

cols_tpsai <- names(efa_data)[c(2:41)]
```

# Exploratory Factor Analysis

Exploratory factor analysis (EFA) is a statistical technique for identifying hidden structure in a set of items. The core idea is simple: if many items correlate with each other, they are probably all measuring some common underlying thing, a latent factor. EFA works by finding the smallest number of latent factors that can reproduce the observed correlations among items. It does this by asking: if these factors exist, what pattern of correlations between items would they produce? The solution that best matches the actual observed correlations, leaving the smallest residuals, is retained. Because EFA makes no assumptions about which items belong to which factor, it is well suited to the early stages of scale development, where the goal is to let the data reveal the underlying structure rather than confirm a predetermined one.

## Factorability

We examing two factorability checks for different sub-scale groupings.

The Kaiser-Meyer-Olkin (KMO) index asks whether the correlations among items are sufficiently compact and patterned to yield reliable factors. Values above .80 indicate good sampling adequacy and values above .90 are excellent. We treat KMO as our primary indicator of factorability.

Bartlett's test of sphericity asks a simpler question: is the correlation matrix significantly different from an identity matrix, where all items would be completely uncorrelated? A significant result (p \< .05) confirms there is at least some correlation structure for factor analysis to work with. However, Bartlett's test is sensitive to sample size and tends to be significant even when correlations are trivially small, so we report it for completeness rather than as a primary decision criterion.

Both tests are computed on the polychoric correlation matrix rather than the standard Pearson correlation matrix, as polychoric correlations are more appropriate for ordinal Likert data [@boateng2018]. A separate matrix is computed for each subscale, since we are conducting independent EFAs for Trust, Distrust, and Corporate Distrust rather than analysing all items together.

```{r}
factorability <- function(R, n) {
  tibble(
    kmo  = KMO(R)$MSA,
    chi2 = cortest.bartlett(R, n)$chisq,
    df   = cortest.bartlett(R, n)$df,
    p    = cortest.bartlett(R, n)$p.value
  )
}

R <- polychoric(efa_data %>% select(all_of(cols_tpsai)))$rho
factorability(R, nrow(efa_data)) %>% kable(digits = 2)
```

## Determining Factors

A key decision in EFA is how many factors to extract. Extracting too few factors leaves meaningful structure unexplained; extracting too many produces factors that are essentially fitting noise in the data.

**Parallel analysis** addresses this by comparing each factor's eigenvalue from the real correlation matrix to the corresponding eigenvalue from random correlation matrices of the same dimensions. Each factor is associated with an *eigenvalue*. Mathematically, an eigenvalue summarises how much of the shared variance among items a factor captures. When EFA extracts a factor, it finds a weighted combination of items that explains as much of the common variance as possible, the eigenvalue is the sum of the squared factor loadings for that factor across all items. A large eigenvalue therefore means many items have high loadings on that factor. Eigenvalues are computed using MINRES estimation, which minimises the residual correlations between the observed and reproduced correlation matrices and makes no distributional assumptions. In parallel analysis, factors are ordered by their eigenvalues from largest to smallest, with the first eigenvalue belonging to the first factor, the one explaining the most variance of the real correlation matrix, the second to the second factor, and so on. For each factor in turn, if the eigenvalue from the real data exceeds the average eigenvalue from the random matrices, meaning the factor is accounting for more variance than a factor extracted from pure noise would, that factor is considered to reflect genuine underlying structure in the data and is worth retaining. The moment the real eigenvalue drops below the random one, which is also what the scree plot visualises, with the blue line (real data) dropping below the red line (random data), that factor and all subsequent ones are considered noise and discarded.

```{r}
fa <- fa.parallel(R,    n.obs = nrow(efa_data), 
            fm = "minres", fa = "fa")
```

Once the number of factors is determined, we extract them using *MINRES* estimation with *oblimin* rotation.

**MINRES** (minimum residual) estimation finds the factor solution that minimises the residuals, the differences between the observed correlation matrix and the one reproduced by the factor model. It makes no distributional assumptions, making it well suited to ordinal Likert data compared to maximum likelihood estimation, which assumes multivariate normality. It is the same method we used to compute eigenvalues and determine the number of factors.

**Rotation** is applied after initial factor extraction to make the solution more interpretable. Without rotation, factors are mathematically defined but often difficult to interpret because most items load moderately onto most factors. Rotation redistributes the variance among factors to achieve a simpler, cleaner pattern where each item loads strongly onto one factor and weakly onto others [@buchanan2014].

We use **oblimin**, which is an oblique rotation method meaning factors are allowed to correlate with one another. This is theoretically appropriate here because items within each subscale are expected to measure related but distinct aspects of the same broader construct:. An orthogonal rotation such as varimax would force factors to be uncorrelated, which is a stronger and less realistic assumption in psychological measurement [@boateng2018]. Importantly, if the factors turn out to be empirically uncorrelated, oblimin will recover the same solution as an orthogonal rotation, so nothing is lost by choosing oblique rotation.

```{r}
efa <- fa(R,
                   nfactors = fa$nfact,
                   n.obs = nrow(efa_data),
                   fm = "minres", rotate = "oblimin")
```

## Evaluating Factors

```{r}
#| fig-width: 10
#| fig-height: 14

plot_loadings <- function(efa, cutoff = 0.15, order = c("theoretical", "empirical")) {
  
  order <- match.arg(order)
  
  loads <- efa$loadings |>
    unclass() |>
    as.data.frame() |>
    rownames_to_column("item")
  
  factor_cols <- setdiff(names(loads), "item")
  
  loads <- loads |>
    rowwise() |>
    mutate(
      primary_factor = factor_cols[which.max(abs(c_across(all_of(factor_cols))))],
      max_loading    = max(abs(c_across(all_of(factor_cols))))
    ) |>
    ungroup()
  
  ordered_items <- if (order == "empirical") {
    loads |>
      arrange(primary_factor, desc(max_loading)) |>
      pull(item)
  } else {
    loads |>
      pull(item)
  }
  
  loads |>
    pivot_longer(cols = -c(item, primary_factor, max_loading), names_to = "factor", values_to = "loading") |>
    mutate(item = factor(item, levels = ordered_items)) |>
    ggplot(aes(x = factor, y = item, fill = loading)) +
    geom_tile() +
    geom_text(aes(label  = round(loading, 2), colour = abs(loading) > 0.5)) +
    scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "black"), guide = "none") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
    scale_y_discrete(limits = rev) +
    theme_minimal() +
    labs(x = "", y = "", fill = "",
         title = if (order == "empirical") "Items ordered by empirical factor" else "Items ordered by theoretical factor")
}

plot_loadings(efa, order = "theoretical")
plot_loadings(efa, order = "empirical")
```

We evaluate factors visually through the heat-map above. We are looking for a factor structure that is robust and simple, that is loadings have high values in one factor only [@buchanan2014].

More specifically, we would like a solution where each factor has at least 3 items and there are no "bad" items that:

-   have a very low primary loading (\< 0.3)

-   have a small difference between primary and secondary loading (\< 0.15)

-   are negatively loaded (since we do not have any reverse-coded item)

If the factor solution present "bad" items, we can address them by:

-   Re-examining the scree-plot of the parallel analysis and inspect whether there is a clear elbow in the plot before the actual data drops below the random one, suggesting that this could be a more suitable smaller number of factors.

-   Removing the items

## Extracting EFA model

Once we are happy with an EFA model, we can extract it for CFA

```{r}
efa_to_cfa(efa)
```


# Inspecting Correlations

## Item-item correlations

First, we examine pairwise inter-item correlations. We flag items correlating above .80 with another item as this indicates the two items share more than 64% of their variance (*r*² = .64). We also visually inspect inter-item correlations for any discontinuity, potentially identifying items worth flagging even if correlated under our pre-set threshold.

We use Polychoric correlations as they are more appropriate for ordinal Likert-scale data than Pearson correlations [@boateng2018].

```{r}
ii_cor_threshold <- 0.8

ii_cor_data <- 
  efa_data %>%
  select(all_of(cols_tpsai)) %>%
  polychoric() %>%.$rho %>%
  as.data.frame() %>%
  rownames_to_column("item1") %>%
  pivot_longer(-item1, names_to = "item2", values_to = "r") %>%
  filter(r > -1 & r < 1) %>%
  filter(item1 < item2)

ii_cor_data %>%
  ggplot(aes(x = r)) +
  geom_histogram(binwidth = 0.01, fill = "steelblue") +
  geom_vline(xintercept = ii_cor_threshold) + geom_vline(xintercept = -ii_cor_threshold) +
  scale_x_continuous(breaks = seq(-1, 1, .2)) +
  coord_cartesian(xlim = c(-1, 1))
```

**Within-factor** high correlations denote items that are moving together as expected.

-   Item pairs with high positive correlations may be asking the same question twice. To reduce the length of our scale, we can consider dropping the weaker item (lower item-total correlation) or merging the pair into a single composite.

-   Item pairs with high negative correlations may indicate that one or more items are worded in the opposite direction to the rest of the sub-scale. If any such items are flagged, we will check for items that may be unintentionally reverse-coded

```{r}
ii_cor_data %>%
  filter(r > ii_cor_threshold) %>% # add in other criteria from visual inspection
  filter(sub("_.*", "", item1) == sub("_.*", "", item2)) %>%
  knitr::kable(digits = 2)
```

**Across-factor inter-item correlations** indicate that two factors are overlapping more than expected. We can inspect these pairs to see if a given item may potentially belong to a different factor than the one we assigned it to.

```{r}
ii_cor_data %>%
  filter(r > ii_cor_threshold) %>%
  filter(sub("_.*", "", item1) != sub("_.*", "", item2)) %>%
  knitr::kable(digits = 2)
```

## Item-total correlation

For each item, we compute the *adjusted item-total correlation*, the correlation between that item and the sum of all other items in the same factor, excluding itself. Items with an adjusted item-total correlation below .30 are flagged for potential removal, as they contribute little to what the factor is seeking to measure [@boateng2018]

As with previous checks, we will also visually inspect the distribution of item-total correlations within each sub-scale, looking for discontinuities that may identify items worth removing even if they fall above our pre-set threshold. Particular attention will be paid to any negatively correlating items, which would suggest an item is moving in the opposite direction to the rest of its sub-scale — a strong signal of potential issues with wording or construct alignment.

```{r}
#| fig-width: 10
#| fig-height: 14

it_cor_threshold <- 0.3

poly_r_drop <- function(items_df) {
  rho <- polychoric(items_df)$rho
  n   <- ncol(rho)
  
  map_dfr(seq_len(n), function(i) {
    others      <- setdiff(seq_len(n), i)
    cov_ir      <- sum(rho[i, others])
    var_rest    <- sum(rho[others, others])
    tibble(
      item    = colnames(rho)[i],
      it_corr = cov_ir / sqrt(var_rest)
    )
  })
}

it_cor_data <- cols_tpsai %>%
  split(sub("_.*", "", .)) %>%
  map_dfr(~ {
    efa_data %>%
      select(all_of(.x)) %>%
      poly_r_drop()
  }, .id = "factor")

it_cor_data %>%
  ggplot(aes(x = it_corr, fill = factor)) +
  facet_wrap(~ factor, ncol = 1, scales = "free_y") +
  geom_histogram(binwidth = 0.05) +
  geom_vline(xintercept =  0) +
  geom_vline(xintercept =  it_cor_threshold, linetype = "dashed") +
  geom_vline(xintercept = -it_cor_threshold, linetype = "dashed") +
  scale_x_continuous(breaks = seq(-1, 1, .2)) +
  coord_cartesian(xlim = c(-1, 1)) +
  labs(x = "", y = "") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none")
```

```{r}
it_cor_items <- it_cor_data %>%
  filter(abs(it_corr) < it_cor_threshold) %>%
  select(item, it_corr)

if (nrow(it_cor_items) > 0) {
  it_cor_items %>% knitr::kable(digits = 2)
}
```

## Factors correlations

We report Spearman correlations between the 8 theoretical factors to examine their inter-relationships. Spearman correlations are used as a rank-based, distribution-free alternative to Pearson, which is more appropriate for composite scores derived from ordinal Likert items [@boateng2018].

```{r}
factors <- names(efa_data)[c(2:41)] %>%
  split(sub("_.*", "", .)) %>%
  imap_dfc(~ efa_data %>%
    transmute(!!.y := rowMeans(pick(all_of(.x)), na.rm = TRUE))
  )

cor(factors, method = "spearman", use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("F1") %>%
  pivot_longer(-F1, names_to = "F2", values_to = "r") %>%
  ggplot(aes(x = F1, y = F2, fill = r)) +
  geom_tile() +
  geom_text(aes(label = round(r, 2)), size = 3.5) +
  scale_fill_gradient2(
    low = "steelblue", mid = "white", high = "firebrick",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(x = "", y = "", fill = "r") +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank())
```

