---
title: "Factor Analysis"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

```{r, echo=FALSE}
library(tidyverse)
library(gtools)
library(psych)
library(knitr)
library(parameters)

set.seed(8675309) # makes sure random numbers are reproducible
```

```{r}
options(readr.show_col_types = FALSE)

source("read_qualtrics_data.R")
d <- load_data("data/pilot/TPSai_17+March+2026_16.59.csv")
cols_tpsai <- d$cols_tpsai

# switch to d$items_data after data collection
efa_data <- read_csv("data/pilot/simulated_efa_data.csv") %>%
  select(all_of(cols_tpsai))
```

# Analysis Structure

Before proceeding to our factor analysis we need to make two key decisions about its structure:

1.  Whether we exclude any items from factor analysis

2.  How to group our sub-scales

## Item Exclusion

Althought we have already covered this step in our item reduction analysis, we re-examine skewness and kurtosis for each item in our new data sample to check whether responses are spread across the scale or piled up at one end. Items where most people gave the same answer have little variation, which makes it harder for factor analysis to detect the underlying structure, regardless of the estimation method used.

Following @kline2016, we treat absolute skewness and kurtosis values above 1.5 as a sign that an item's responses are too bunched together to be useful. Items exceeding this threshold will be considered for removal alongside the other item reduction criteria. Additionally, we may consider keeping items with high skew or kurtosis for comparison or reporting purposes but flag them for removal from factor analysis.

```{r}
dist_threshold <- 1.5

dist_data <- efa_data %>%
  summarise(
    across(all_of(everything()), list(
        skew = ~ psych::skew(.x,    na.rm = TRUE),
        kurt = ~ psych::kurtosi(.x, na.rm = TRUE)
      ), .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("subscale", "q_n"), sep = "_", remove = FALSE)

ggplot(dist_data, aes(x = skew, y = kurt, colour = subscale, label = item)) +
  geom_point() +
  geom_hline(yintercept = c(-dist_threshold, dist_threshold), linetype = "dashed") +
  geom_vline(xintercept = c(-dist_threshold, dist_threshold), linetype = "dashed") +
  labs(x = "Skewness", y = "Kurtosis")
```

```{r}
dist_data %>%
  filter(abs(skew) > dist_threshold | abs(kurt) > dist_threshold) %>% 
  select(subscale, item, skew, kurt) %>% 
  knitr::kable(digits = 2)
```

## Sub-scale Grouping

We consider three options for factor analysis:

1.  Conduct separate factor analysis for each sub-scale

2.  Conduct a single factor analysis across sub-scales

3.  Conduct a factor analysis for trust and distrust items, merging Trust and Corporate Distrust into a single sub-scale

Before empirical examination, some argument worth considering are:

-   the theoretical independence of Trust and Distrust as related but distinct constructs warrants separate investigation [@mcknight2001; @scharowski24], suggesting either option 1 or 3.

-   combining all items in a single analysis would introduce a mixed-valence structure. Trust items are positively worded while Distrust and Corporate Distrust items are negatively worded, which risks producing factors driven by wording artefacts rather than substantive constructs [@zhang2016], suggesting either option 1 or 3.

-   Corporate Distrust is conceptually distinct from Distrust as it targets an organisational entity rather than an AI system directly, suggesting option 1 over option 3, although weakly because it could be a factor.

-   Trust and Distrust (including Corporate Distrust) items are more or less of equal length (22 vs. 21) suggesting two length-balanced sub-scales, suggesting option 3 over 1.

### Correlations

First, we report Spearman correlations between the three sub-scale mean scores (Trust, Distrust, and Corporate Distrust) to examine their inter-relationships prior to factor analysis. Spearman correlations are used as a rank-based, distribution-free alternative to Pearson, which is more appropriate for composite scores derived from ordinal Likert items [@boateng2018].

If any two subscales correlate above .80, indicating substantial construct overlap, we consider combining them for factor analysis into either option 2 or 3.

```{r}
subscales <- efa_data %>%
  mutate(
    Trust    = rowMeans(pick(all_of(d$cols_trust)),    na.rm = TRUE),
    Distrust = rowMeans(pick(all_of(d$cols_distrust)), na.rm = TRUE),
    "Corporate Distrust"  = rowMeans(pick(all_of(d$cols_company)),  na.rm = TRUE)
  ) %>%
  select(Trust, Distrust, "Corporate Distrust")


cor_subscales <- cor(subscales, method = "spearman", use = "pairwise.complete.obs")

cor_subscales %>% 
  as.data.frame() %>% 
  rownames_to_column("S1") %>% 
  pivot_longer(-S1, names_to = "S2", values_to = "r") %>% 
  filter(S1 < S2) %>% 
  kable(digits = 2)
```

### Factorability

We examing two factorability checks for different sub-scale groupings.

The Kaiser-Meyer-Olkin (KMO) index asks whether the correlations among items are sufficiently compact and patterned to yield reliable factors. Values above .80 indicate good sampling adequacy and values above .90 are excellent. We treat KMO as our primary indicator of factorability.

Bartlett's test of sphericity asks a simpler question: is the correlation matrix significantly different from an identity matrix, where all items would be completely uncorrelated? A significant result (p \< .05) confirms there is at least some correlation structure for factor analysis to work with. However, Bartlett's test is sensitive to sample size and tends to be significant even when correlations are trivially small, so we report it for completeness rather than as a primary decision criterion.

Both tests are computed on the polychoric correlation matrix rather than the standard Pearson correlation matrix, as polychoric correlations are more appropriate for ordinal Likert data [@boateng2018]. A separate matrix is computed for each subscale, since we are conducting independent EFAs for Trust, Distrust, and Corporate Distrust rather than analysing all items together.

```{r}
factorability <- function(R, n) {
  tibble(
    kmo  = KMO(R)$MSA,
    chi2 = cortest.bartlett(R, n)$chisq,
    df   = cortest.bartlett(R, n)$df,
    p    = cortest.bartlett(R, n)$p.value
  )
}

R_all        <- polychoric(efa_data %>% select(all_of(d$cols_tpsai)))$rho
R_trust      <- polychoric(efa_data %>% select(all_of(d$cols_trust)))$rho
R_distrust_1 <- polychoric(efa_data %>% select(all_of(d$cols_distrust)))$rho
R_distrust_2 <- polychoric(efa_data %>% select(all_of(c(d$cols_distrust, d$cols_company))))$rho
R_company    <- polychoric(efa_data %>% select(all_of(d$cols_company)))$rho

bind_rows(
  factorability(R_all,    nrow(efa_data)),
  factorability(R_trust,    nrow(efa_data)),
  factorability(R_distrust_1, nrow(efa_data)),
  factorability(R_company,  nrow(efa_data)),
  factorability(R_distrust_2, nrow(efa_data)),
) %>%
  mutate(subscale = c("All", "Trust", "Distrust Single", "Corporate Distrust", "Distrust (including Corporate)")) %>%
  select(subscale, everything()) %>%
  kable(digits = 3,
        col.names = c("Subscale", "KMO", "χ²", "df", "p"))
```

# Exploratory Factor Analysis

Exploratory factor analysis (EFA) is a statistical technique for identifying hidden structure in a set of items. The core idea is simple: if many items correlate with each other, they are probably all measuring some common underlying thing, a latent factor. EFA works by finding the smallest number of latent factors that can reproduce the observed correlations among items. It does this by asking: if these factors exist, what pattern of correlations between items would they produce? The solution that best matches the actual observed correlations, leaving the smallest residuals, is retained. Because EFA makes no assumptions about which items belong to which factor, it is well suited to the early stages of scale development, where the goal is to let the data reveal the underlying structure rather than confirm a predetermined one.

## Determining Factors

A key decision in EFA is how many factors to extract. Extracting too few factors leaves meaningful structure unexplained; extracting too many produces factors that are essentially fitting noise in the data.

**Parallel analysis** addresses this by comparing each factor's eigenvalue from the real correlation matrix to the corresponding eigenvalue from random correlation matrices of the same dimensions. Each factor is associated with an *eigenvalue*. Mathematically, an eigenvalue summarises how much of the shared variance among items a factor captures. When EFA extracts a factor, it finds a weighted combination of items that explains as much of the common variance as possible, the eigenvalue is the sum of the squared factor loadings for that factor across all items. A large eigenvalue therefore means many items have high loadings on that factor. Eigenvalues are computed using MINRES estimation, which minimises the residual correlations between the observed and reproduced correlation matrices and makes no distributional assumptions. In parallel analysis, factors are ordered by their eigenvalues from largest to smallest, with the first eigenvalue belonging to the first factor, the one explaining the most variance of the real correlation matrix, the second to the second factor, and so on. For each factor in turn, if the eigenvalue from the real data exceeds the average eigenvalue from the random matrices, meaning the factor is accounting for more variance than a factor extracted from pure noise would, that factor is considered to reflect genuine underlying structure in the data and is worth retaining. The moment the real eigenvalue drops below the random one, which is also what the scree plot visualises, with the blue line (real data) dropping below the red line (random data), that factor and all subsequent ones are considered noise and discarded.

```{r}
fa_trust <- fa.parallel(R_trust,    n.obs = nrow(efa_data), 
            fm = "minres", fa = "fa",
            main = "Parallel Analysis — Trust")

fa_distrust <- fa.parallel(R_distrust_2, n.obs = nrow(efa_data), 
            fm = "minres", fa = "fa",
            main = "Parallel Analysis — Distrust")
```

Once the number of factors is determined, we extract them using *MINRES* estimation with *oblimin* rotation.

**MINRES** (minimum residual) estimation finds the factor solution that minimises the residuals, the differences between the observed correlation matrix and the one reproduced by the factor model. It makes no distributional assumptions, making it well suited to ordinal Likert data compared to maximum likelihood estimation, which assumes multivariate normality. It is the same method we used to compute eigenvalues and determine the number of factors.

**Rotation** is applied after initial factor extraction to make the solution more interpretable. Without rotation, factors are mathematically defined but often difficult to interpret because most items load moderately onto most factors. Rotation redistributes the variance among factors to achieve a simpler, cleaner pattern where each item loads strongly onto one factor and weakly onto others [@buchanan2014].

We use **oblimin**, which is an oblique rotation method meaning factors are allowed to correlate with one another. This is theoretically appropriate here because items within each subscale are expected to measure related but distinct aspects of the same broader construct:. An orthogonal rotation such as varimax would force factors to be uncorrelated, which is a stronger and less realistic assumption in psychological measurement [@boateng2018]. Importantly, if the factors turn out to be empirically uncorrelated, oblimin will recover the same solution as an orthogonal rotation, so nothing is lost by choosing oblique rotation.

```{r}
efa_trust    <- fa(R_trust,
                   nfactors = 2,
                   n.obs = nrow(efa_data),
                   fm = "minres", rotate = "oblimin")

efa_distrust <- fa(R_distrust_2,
                   nfactors = 2,
                   n.obs = nrow(efa_data),
                   fm = "minres", rotate = "oblimin")
```

## Evaluating Factors

```{r}
plot_loadings <- function(efa, cutoff = 0.15) {
  
  loads <- efa$loadings |>
    unclass() |>
    as.data.frame() |>
    rownames_to_column("item")
  
  factor_cols <- setdiff(names(loads), "item")
  
  ordered_items <- loads |>
    rowwise() |>
    mutate(
      primary_factor = factor_cols[which.max(abs(c_across(all_of(factor_cols))))],
      max_loading    = max(abs(c_across(all_of(factor_cols))))
    ) |>
    ungroup() |>
    arrange(primary_factor, desc(max_loading)) |>
    pull(item)
  
  loads |>
    pivot_longer(cols = -item, names_to = "factor", values_to = "loading") |>
    mutate(item = factor(item, levels = ordered_items)) |>
    ggplot(aes(x = factor, y = item, fill = loading)) +
    geom_tile() +
    geom_text(aes(
      label  = round(loading, 2),
      colour = abs(loading) > 0.5
    )) +
    scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "black"), guide = "none") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
    scale_y_discrete(limits = rev) +
    theme_minimal() +
    labs(x = "Factor", y = "Item", fill = "Loading")
}

plot_loadings(efa_trust)
plot_loadings(efa_distrust)
```

We evaluate factors visually through the heat-map above. We are looking for a factor structure that is robust and simple, that is loadings have high values in one factor only [@buchanan2014].

More specifically, we would like a solution where each factor has at least 3 items and there are no "bad" items that:

-   have a very low primary loading (\< 0.3)

-   have a small difference between primary and secondary loading (\< 0.15)

-   are negatively loaded (since we do not have any reverse-coded item)

If the factor solution present "bad" items, we can address them by:

-   Re-examining the scree-plot of the parallel analysis and inspect whether there is a clear elbow in the plot before the actual data drops below the random one, suggesting that this could be a more suitable smaller number of factors.

-   Removing the items

## Extracting EFA model

Once we are happy with an EFA model, we can extract it for CFA

```{r}
efa_to_cfa(efa_trust)
```

```{r}
efa_to_cfa(efa_distrust)
```
