---
title: "Confirmatory Factor Analysis (CFA)"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

# Data Processing

```{r}
library(tidyverse)
library(psych)
library(lavaan)
library(semTools)
library(knitr)

set.seed(8675309)
```

```{r}
source("read_qualtrics_data.R")
d <- load_data("data/pilot/TPSai_17+March+2026_16.59.csv")
cols_tpsai <- d$cols_tpsai

# switch to d$items_data after data collection
cfa_data <- read_csv("data/pilot/simulated_cfa_data.csv") %>%
  select(all_of(cols_tpsai))
```

# Confirmatory Factor Analysis

Confirmatory factor analysis (CFA) tests whether a hypothesised factor structure — derived from theory and/or EFA results — replicates in the data. Unlike EFA, which lets the data reveal structure freely, CFA is a hypothesis-testing procedure: we specify in advance which items load onto which factors and which cross-loadings are constrained to zero, then evaluate how well that model fits the observed covariance matrix.

We use the **WLSMV** (weighted least squares, mean- and variance-adjusted) estimator throughout. WLSMV is the recommended estimator for ordinal Likert data because it treats items as ordered-categorical rather than continuous, avoiding the underestimation of factor loadings and the distorted fit indices that arise when standard maximum-likelihood estimation is applied to non-normal ordinal responses.

## Build Models

```{r}
trust_1f_model = cfa(paste("TR =~", paste(d$cols_trust, collapse = " + ")), 
                     cfa_data, estimator = "WLSMV", ordered = TRUE)


```

## Model Fit

We evaluate fit using the indices recommended by @boateng2018 and @briones2026. No single index is sufficient; good fit requires convergent evidence across indices.

| Index | Good fit | Acceptable | Poor fit |
|-------|----------|------------|----------|
| CFI   | ≥ .95    | .90–.94    | \< .90   |
| TLI   | ≥ .95    | .90–.94    | \< .90   |
| RMSEA | ≤ .06    | .06–.08    | \> .10   |
| SRMR  | ≤ .06    | .06–.08    | \> .08   |

```{r}
fitMeasures(trust_1f_model)
```

```{r}
extract_fit <- function(fit, subscale, model_label) {
  fi <- fitMeasures(fit, c("chisq", "df", "pvalue",
                            "cfi", "tli",
                            "rmsea", "rmsea.ci.lower", "rmsea.ci.upper",
                            "srmr"))
  tibble(
    Subscale = subscale,
    Model    = model_label,
    `χ²`    = round(fi["chisq"], 2),
    df       = fi["df"],
    p        = round(fi["pvalue"], 3),
    CFI      = round(fi["cfi"],   3),
    TLI      = round(fi["tli"],   3),
    RMSEA    = round(fi["rmsea"], 3),
    `90% CI` = paste0("[", round(fi["rmsea.ci.lower"], 3),
                      ", ", round(fi["rmsea.ci.upper"], 3), "]"),
    SRMR     = round(fi["srmr"],  3)
  )
}

bind_rows(
  extract_fit(trust_1f_model, "Trust", "One factor"),
) %>%
  kable()
```

## Model Comparison

When both models exist we compare them with a scaled chi-square difference test. Because WLSMV statistics are not ordinary chi-squares, `lavTestLRT()` applies the Satorra-Bentler correction appropriate for the difference test [@satorra2010].

We prefer the EFA-derived model when **both** criteria are met: the difference test is significant (p \< .05) **and** ΔCFI \> .01. Using both criteria guards against retaining a complex model purely on the basis of a statistically significant but practically trivial improvement in fit [@chen2007].

```{r}
compare_models <- function(fits, label) {

  if (is.null(fits$efa_derived)) {
    return(tibble(Subscale = label, `Δχ²` = NA, `Δdf` = NA, p = NA,
                  ΔCFI = NA, ΔRMSEA = NA,
                  Decision = "No comparison — single-factor solution only."))
  }

  lrt <- lavTestLRT(fits$one_factor, fits$efa_derived,
                    method = "satorra.bentler.2010",
                    model.names = c("One-factor", "EFA-derived"))

  delta_cfi   <- round(fitMeasures(fits$efa_derived, "cfi") -
                       fitMeasures(fits$one_factor,  "cfi"), 3)
  delta_rmsea <- round(fitMeasures(fits$efa_derived, "rmsea") -
                       fitMeasures(fits$one_factor,  "rmsea"), 3)
  p_val       <- lrt[2, "Pr(>Chisq)"]

  tibble(
    Subscale  = label,
    `Δχ²`    = round(lrt[2, "Chisq diff"], 2),
    `Δdf`    = lrt[2, "Df diff"],
    p         = round(p_val, 3),
    ΔCFI      = delta_cfi,
    ΔRMSEA    = delta_rmsea,
    Decision  = case_when(
      p_val < .05 & delta_cfi > .01 ~
        "Prefer EFA-derived (sig. Δχ² and ΔCFI > .01)",
      p_val < .05 ~
        "Sig. Δχ² but ΔCFI ≤ .01 — prefer one-factor (simpler)",
      TRUE ~
        "Non-sig. Δχ² — prefer one-factor (simpler)"
    )
  )
}

bind_rows(
  compare_models(fits_trust,    "Trust"),
  compare_models(fits_distrust, "Distrust"),
  # compare_models(fits_company,  "Corporate Distrust")
) %>%
  kable()
```
