---
title: "Confirmatory Factor Analysis (CFA)"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

```{r, echo=FALSE}
library(tidyverse)
library(psych)
library(knitr)
library(lavaan)
library(semTools)

seed <- 2641388
set.seed(seed) # makes sure random numbers are reproducible
```

```{r}
options(readr.show_col_types = FALSE)

source("simulate_data.R")

cfa_data <- simulate_data(
  sample_path = "data/pilot/TPSai_29+March+2026_15.56.csv", 
  factors_path =  "data/factors.csv",
  seed = seed,
  n = 300)

cols_tpsai <- names(cfa_data)[c(2:41)]
```

Confirmatory factor analysis (CFA) tests whether a hypothesised factor structure — derived from theory and/or EFA results — replicates in the data. Unlike EFA, which lets the data reveal structure freely, CFA is a hypothesis-testing procedure: we specify in advance which items load onto which factors and which cross-loadings are constrained to zero, then evaluate how well that model fits the observed covariance matrix.

We use the **WLSMV** (weighted least squares, mean- and variance-adjusted) estimator throughout. WLSMV is the recommended estimator for ordinal Likert data because it treats items as ordered-categorical rather than continuous, avoiding the underestimation of factor loadings and the distorted fit indices that arise when standard maximum-likelihood estimation is applied to non-normal ordinal responses.

# Build Models

```{r}
model_1F <- paste("MR1 =~", paste(cols_tpsai, collapse = " + "))

model_EFA <- '
MR2 =~ COMP_01 + COMP_02 + COMP_03 + CRIT_02 + HUMN_01 + HUMN_04 + PERS_01 + PERS_03 + PERS_04 + RISK_01 + SAFE_01 + SAFE_03
MR1 =~ CORP_01 + CORP_03 + CORP_05 + CRIT_03 + CRIT_04 + HUMN_03 + HUMN_05 + RISK_02 + RISK_04 + SAFE_04
MR3 =~ CRIT_01 + CRIT_05 + EMOT_01 + EMOT_02 + EMOT_05 + PERS_02 + PERS_05 + SAFE_02
MR4 =~ COMP_05 + CORP_02 + EMOT_03 + RISK_03
MR5 =~ COMP_04 + CORP_04 + EMOT_04 + HUMN_02 + RISK_05 + SAFE_05
'
```

```{r}
cfa_model_1F <- cfa(model_1F, cfa_data, estimator = "WLSMV", ordered = TRUE)

cfa_model_EFA <- cfa(model_EFA, cfa_data, estimator = "WLSMV", ordered = TRUE)
```

# Model Fit

We evaluate fit using the indices recommended by @boateng2018 and @briones2026. No single index is sufficient; good fit requires convergent evidence across indices.

| Index | Good fit | Acceptable | Poor fit |
|-------|----------|------------|----------|
| CFI   | ≥ .95    | .90–.94    | \< .90   |
| TLI   | ≥ .95    | .90–.94    | \< .90   |
| RMSEA | ≤ .06    | .06–.08    | \> .10   |
| SRMR  | ≤ .06    | .06–.08    | \> .08   |

```{r}
extract_fit <- function(fit, model_label) {
  fi <- fitMeasures(fit, c("chisq", "df", "pvalue",
                            "cfi", "tli",
                            "rmsea", "rmsea.ci.lower", "rmsea.ci.upper",
                            "srmr"))
  tibble(
    Model    = model_label,
    `χ²`    = round(fi["chisq"], 2),
    df       = fi["df"],
    # p        = round(fi["pvalue"], 3),
    CFI      = round(fi["cfi"],   3),
    TLI      = round(fi["tli"],   3),
    RMSEA    = round(fi["rmsea"], 3),
    `90% CI` = paste0("[", round(fi["rmsea.ci.lower"], 3),
                      ", ", round(fi["rmsea.ci.upper"], 3), "]"),
    SRMR     = round(fi["srmr"],  3)
  )
}

bind_rows(
  extract_fit(cfa_model_1F, "1F"),
  extract_fit(cfa_model_EFA, "EFA"),
) %>%
  kable()
```

We can also compare models with a scaled chi-square difference test. We prefer more complex models when **both** criteria are met: the difference test is significant (p \< .05) and ΔCFI \> .01. Using both criteria guards against retaining a complex model purely on the basis of a statistically significant but practically trivial improvement in fit.

```{r}
lavTestLRT(cfa_model_1F, cfa_model_EFA, model.names = c("1F", "EFA"))
```
