---
title: "Confirmatory Factor Analysis (CFA)"
author: "Riccardo Volpato"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

```{r}
library(tidyverse)
library(psych)
library(knitr)
library(lavaan)
library(semTools)

set.seed(8675309)
```

```{r}
source("read_qualtrics_data.R")
d <- load_data("data/pilot/TPSai_17+March+2026_16.59.csv")
cols_tpsai <- d$cols_tpsai

# switch to d$items_data after data collection
cfa_data <- read_csv("data/pilot/simulated_cfa_data.csv") %>%
  select(all_of(cols_tpsai))
```

Confirmatory factor analysis (CFA) tests whether a hypothesised factor structure — derived from theory and/or EFA results — replicates in the data. Unlike EFA, which lets the data reveal structure freely, CFA is a hypothesis-testing procedure: we specify in advance which items load onto which factors and which cross-loadings are constrained to zero, then evaluate how well that model fits the observed covariance matrix.

We use the **WLSMV** (weighted least squares, mean- and variance-adjusted) estimator throughout. WLSMV is the recommended estimator for ordinal Likert data because it treats items as ordered-categorical rather than continuous, avoiding the underestimation of factor loadings and the distorted fit indices that arise when standard maximum-likelihood estimation is applied to non-normal ordinal responses.

# Build Models

```{r}
trust_1f_model <- cfa(paste("TR =~", paste(d$cols_trust, collapse = " + ")),
                      cfa_data, estimator = "WLSMV", ordered = TRUE)

distrust_1f_model <- cfa(paste("DT =~", paste(d$cols_distrust, collapse = " + "), " + ",
                               paste(d$cols_company, collapse = " + ")),
                      cfa_data, estimator = "WLSMV", ordered = TRUE)
```

```{r}
trust_efa_eq <- '
  MR1 =~ TR_02 + TR_04 + TR_06 + TR_09 + TR_10 + TR_12 + TR_13 + TR_14 + TR_15 + TR_16 + TR_18 + TR_20 + TR_22
  MR2 =~ TR_01 + TR_03 + TR_05 + TR_07 + TR_08 + TR_11 + TR_17 + TR_19 + TR_21
'
trust_efa_model <- cfa(trust_efa_eq,
                      cfa_data, estimator = "WLSMV", ordered = TRUE)

distrust_efa_eq <- '
MR1 =~ DT_01 + DT_02 + DT_04 + DT_06 + DT_10 + DT_11 + DT_13 + DT_14 + CD_04 + CD_05
MR2 =~ DT_03 + DT_05 + DT_07 + DT_08 + DT_09 + DT_12 + CD_01 + CD_02 + CD_03 + CD_06 + CD_07
'
distrust_efa_model <- cfa(distrust_efa_eq,
                        cfa_data, estimator = "WLSMV", ordered = TRUE)
```

# Model Fit

We evaluate fit using the indices recommended by @boateng2018 and @briones2026. No single index is sufficient; good fit requires convergent evidence across indices.

| Index | Good fit | Acceptable | Poor fit |
|-------|----------|------------|----------|
| CFI   | ≥ .95    | .90–.94    | \< .90   |
| TLI   | ≥ .95    | .90–.94    | \< .90   |
| RMSEA | ≤ .06    | .06–.08    | \> .10   |
| SRMR  | ≤ .06    | .06–.08    | \> .08   |

```{r}
extract_fit <- function(fit, subscale, model_label) {
  fi <- fitMeasures(fit, c("chisq", "df", "pvalue",
                            "cfi", "tli",
                            "rmsea", "rmsea.ci.lower", "rmsea.ci.upper",
                            "srmr"))
  tibble(
    Subscale = subscale,
    Model    = model_label,
    `χ²`    = round(fi["chisq"], 2),
    df       = fi["df"],
    # p        = round(fi["pvalue"], 3),
    CFI      = round(fi["cfi"],   3),
    TLI      = round(fi["tli"],   3),
    RMSEA    = round(fi["rmsea"], 3),
    `90% CI` = paste0("[", round(fi["rmsea.ci.lower"], 3),
                      ", ", round(fi["rmsea.ci.upper"], 3), "]"),
    SRMR     = round(fi["srmr"],  3)
  )
}

bind_rows(
  extract_fit(trust_1f_model, "Trust", "1F"),
  extract_fit(trust_efa_model, "Trust", "EFA"),
  extract_fit(distrust_1f_model, "Disrust", "1F"),
  extract_fit(distrust_efa_model, "Disrust", "EFA"),
) %>%
  kable()
```

We can also compare models with a scaled chi-square difference test.

We prefer more complex models when **both** criteria are met: the difference test is significant (p \< .05) and ΔCFI \> .01. Using both criteria guards against retaining a complex model purely on the basis of a statistically significant but practically trivial improvement in fit.

```{r}
lavTestLRT(trust_1f_model, trust_efa_model, model.names = c("1F", "EFA"))
```

```{r}
lavTestLRT(distrust_1f_model, distrust_efa_model, model.names = c("1F", "EFA"))
```
