---
title: "Pilot Evaluation"
author: "Riccardo Volpato"
format: html
editor: visual
bibliography: references.bib
execute:
  message: false
  warning: false
---

# TPSai: Expert evaluation

## Data Processing

```{r}
library(tidyverse)
library(gtools)
library(psych)
library(irr)
library(knitr)

set.seed(8675309) # makes sure random numbers are reproducible
```

```{r}
# read Qualtrics data without metadata rows
file_path <- "data/pilot/TPSai+-+Experts_11+March+2026_12.08.csv"
raw_data <- read_csv(file_path, n_max = 2)
colnames <- names(raw_data)
raw_data <- read_csv(file_path, col_names = colnames, skip = 3)
```

```{r}
# define variables for questions and answers
cols_trust    <- paste0("TR_", 1:23)
cols_distrust <- paste0("DT_", 1:13)
cols_company  <- paste0("CD_", 1:7)
cols_tpsai <- c(cols_trust, cols_distrust, cols_company)
choices <- c("Keep" = 1, "Drop" = 2, "Reword" = 3)
```

```{r}
# generate random data for code development before data collection
random_data <- as.data.frame(
  lapply(cols_tpsai, function(col) {
    sample(names(choices), size=5, replace=TRUE)
  })
)
colnames(random_data) <- cols_tpsai
```

```{r}
analysis_data <- random_data # switch to raw_data after data collection

experts_data <- analysis_data %>%
  select(all_of(cols_tpsai))  %>%
  mutate(across(all_of(cols_tpsai), ~ choices[.x]))
```

## Experts Evaluation: Inter-Rater Agreement

```{r}
ratings_matrix <- experts_data %>%
  select(all_of(cols_tpsai)) %>%
  mutate(across(everything(), as.integer)) %>%
  as.matrix()

overall_alpha <- kripp.alpha(ratings_matrix, method = "nominal")
overall_alpha

interpret_alpha <- case_when(
  overall_alpha$value < 0.20 ~ "Slight agreement",
  overall_alpha$value < 0.40 ~ "Fair agreement",
  overall_alpha$value < 0.60 ~ "Moderate agreement",
  overall_alpha$value < 0.80 ~ "Substantial agreement",
  TRUE                        ~ "Almost perfect agreement"
)

cat("Krippendorff's alpha =", round(overall_alpha$value, 3), "-", interpret_alpha)
```

Expert judges evaluated each item and voted to Keep, Drop, or Reword it. The majority vote determined the final recommendation for each item. When two options tied, the following rules resolved the decision.

1.  Keep and Reword ties favoured Reword on the basis that any doubt about an item warrants revision rather than acceptance.
2.  Drop and Reword ties favoured Drop on the basis that if an item is difficult enough to reword that experts are split, it is better removed.
3.  Keep and Drop ties favoured Keep, giving the item the benefit of the doubt for further evaluation in subsequent validation stages.

```{r}
decisions <- experts_data %>%
  pivot_longer(cols = all_of(cols_tpsai), names_to = "item", values_to = "rating") %>%
  mutate(rating = names(choices)[rating]) %>%
  group_by(item) %>%
  summarise(
    n_keep    = sum(rating == "Keep"),
    n_drop    = sum(rating == "Drop"),
    n_reword  = sum(rating == "Reword"),
    pct_agree = round(max(table(rating)) / n(), 2),
    .groups   = "drop"
  ) %>%
  mutate(
    decision = case_when(
      n_reword > n_keep  & n_reword > n_drop  ~ "Reword",
      n_drop   > n_keep  & n_drop   > n_reword ~ "Drop",
      n_keep   > n_drop  & n_keep   > n_reword ~ "Keep",
      n_keep == n_reword & n_keep > n_drop    ~ "Reword",
      n_drop == n_reword & n_drop > n_keep    ~ "Drop",
      n_keep == n_drop  & n_keep > n_reword   ~ "Keep"
    ),
    item = factor(item, levels = cols_tpsai)
  ) %>%
  arrange(item)

kable(decisions, col.names = c("Item", "Keep", "Drop", "Reword", "% Agreement", "Decision"))
```
