---
title: "Pilot Evaluation"
author: "Riccardo Volpato"
format: html
editor: visual
bibliography: references.bib
---

# TPSai: Quantitative Pilot Evaluation

## Data Processing

First, let's load the required packages

```{r}
# packages needed
library(tidyverse)
library(gtools)
library(psych)
library(faux)

set.seed(8675309) # makes sure random numbers are reproducible
```

Then, read and process the data downloaded from Qualtrics.

```{r}

raw_pilot_data <- read_csv("data/pilot/quant_pilot_data.csv", n_max = 2)
colnames <- names(raw_pilot_data)
raw_pilot_data <- read_csv("data/pilot/quant_pilot_data.csv", col_names = colnames, skip = 3)

# simulate 
params <- list(
  list(prob = c(
    "Strongly disagree" = 10,
    "Disagree" = 30,
    "Neither agree nor disagree" = 40,
    "Agree" = 20,
    "Strongly agree" = 5
  ))
) |> rep(6)
names(params) <- paste0("ppta_", 1:6)


ppt_a <- faux::rmulti(n = 100, 
                      dist = c(
                        ppta_1 = "likert",
                        ppta_2 = "likert",
                        ppta_3 = "likert",
                        ppta_4 = "likert",
                        ppta_5 = "likert",
                        ppta_6 = "likert"
                      ),
                      params = params, 
                      r = .5)

# define column ranges
cols_ppt_a    <- paste0("PPT-A_", 1:6)
cols_trust    <- paste0("TRUST_", 1:23)
cols_distrust <- paste0("DISTRUST_", 1:13)
cols_company  <- paste0("CORP-DIS_", 1:7)
cols_tpsai <- c(cols_trust, cols_distrust, cols_company)
cols_likert <- c(cols_tpsai, cols_ppt_a)

# map likert levels to numerical values
likert_levels <- c(
  "Strongly disagree" = 1,
  "Disagree" = 2,
  "Neither agree nor disagree" = 3,
  "Agree" = 4,
  "Strongly agree" = 5,
  "Don't know" = NA
)



# load and process data
pilot_csv <- raw_pilot_data %>%
  # remove metadata row and question text row
  filter(!str_detect(Status, "ImportId|Response Type")) %>%
  # select only columns we plan to use
  select(FREQ, `PLAT-PSY`, `PLAT-GEN`, SINGLE_1:SINGLE_2, all_of(cols_likert)) %>%
  mutate(across(all_of(cols_likert), ~ likert_levels[.x])) %>%
  # simulate one column with >.3 missingness
  faux::messy(prop = .4, "TRUST_1") %>%
  # make a high correlation between DISTRUST_1 and _2
  mutate(DISTRUST_2 = DISTRUST_1 + sample(-1:1, n(), replace = TRUE),
         DISTRUST_2 = case_match(DISTRUST_2, 0 ~ 1, 6 ~ 5, .default = DISTRUST_2))
```

## Item Evaluation

Let's quantitatively evaluate the TPSai items to identify which items might be unsuitable for scale development.

### High missingness

First, let's search for items across all TPSai sub-scales that many participants responded "*Don't know*" to. High missingness items likely capture constructs that participants are highly uncertain about and can reduce the effective sample size and bias item statistics like means and correlations. Given that we have a substantial number of survey items (23 trust, 13 distrust, and 6 company) removing high missingness items is a way of shortening our survey by focusing on constructs that people have clearer ideas about.

We decided to drop items exceeding 30% missing responses. Our rationale is that if at least 3 out of 10 individuals were not able to meaningfully engage with an item, dropping it simplifies our scale and allows it to focus on measuring items with generally clearer responses. 30% is a relatively moderate threshold for removal of items with high missingness, which might avoid premature dropping of items with moderate missingness may still yield interpretable results.

Nonetheless, following the notion of midpoint abuse as pseudo-missingness @nadler2015, whereby respondents default to the neutral option rather than express a genuine attitude, often due to item ambiguity or poor wording, we also drop items where both missing and neutral responses ("Neither agree nor disagree") each exceeded 15%. The combination of these two patterns suggests that a substantial proportion of respondents were unable or unwilling to engage meaningfully with the item.

```{r}
missing_items <- pilot_csv %>%
  summarise(
    across(all_of(cols_tpsai), ~ mean(is.na(.x)), .names = "{.col}_missing")
  ) %>%
  pivot_longer(
    everything(),
    names_to  = c("item", ".value"),
    names_sep = "_(?=[^_]+$)"
  ) %>%
  separate(item, c("scale", "q_n"), sep = "_", remove = FALSE)

ggplot(missing_items, aes(x = missing, fill = scale)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = .3)
```


```{r}
items_to_drop <- missing_items %>%
  filter(missing > .3) %>%  # add in other criteria from visual inspection
  pull(item)
items_to_drop
```

```{r}
# dropping items
cols_trust    <- setdiff(cols_trust,    items_to_drop)
cols_distrust <- setdiff(cols_distrust, items_to_drop)
cols_company  <- setdiff(cols_company,  items_to_drop)
cols_tpsai    <- setdiff(cols_tpsai,    items_to_drop)
cols_likert   <- setdiff(cols_likert,   items_to_drop)

pilot_csv <- pilot_csv %>%
  select(-all_of(items_to_drop))
```

### Item correlations

Next, we examine pairwise polychoric inter-item correlations within each TPSai subscale (Trust, Distrust, and AI Company) to identify redundant items. Polychoric correlations are used as they are more appropriate for ordinal Likert-scale data than Pearson correlations @boateng2018. Items correlating above .80 with another item within the same subscale are considered redundant, as this indicates the two items share more than 64% of their variance (*r*² = .64), leaving little unique contribution to the scale.

```{r}
ii_cor_threshold <- 0.8

ii_cor_data <- list(
  Trust    = cols_trust,
  Distrust = cols_distrust,
  Company  = cols_company
) %>%
  map_dfr(~ {
    pilot_csv %>%
      select(all_of(.x)) %>%
      polychoric() %>%
      .$rho %>%
      as.data.frame() %>%
      rownames_to_column("item1") %>%
      pivot_longer(-item1, names_to = "item2", values_to = "r")
  }, .id = "subscale")

ii_cor_data %>%
  filter(r < 1) %>%
  ggplot(aes(x = r, fill= subscale)) +
  facet_wrap(~subscale, ncol = 1, scales = "free_y") +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = ii_cor_threshold) +
  geom_vline(xintercept = -ii_cor_threshold) +
  scale_x_continuous(breaks = seq(-1, 1, .2), limits = c(-1, 1))
```

```{r}
too_correlated <- ii_cor_data %>%
  filter(item1 < item2, r > ii_cor_threshold)
```


## Distributions of all questions

Just for info, no removal of low variability

```{r}

```


## Explore Individual Bias

To account for general tendencies to trust automated technologies, we have include the Propensity to Trust in Automated Technology (PTT-A) scale @scholz2025 in the survey battery (6-items short version). PTT-A response provide an individual baseline measure of general trust toward automated systems, which may influence their ratings of psychological support from generative AI.

By correlating PTT-A scores with the subscales of Trust, Distrust, and AI Company, we can investigate whether TPSai scales capture context-specific trust constructs distinct from general tendencies to trust automated systems. Low-to-moderate correlations would indicate that our scale measures unique psychological attitudes toward generative AI for psychological support, rather than merely reflecting general trust dispositions.

```{r}
# Reverse score PTT-A items 3 and 4 (scale: 1-5)
pilot_csv <- pilot_csv %>%
  mutate(
    `PPT-A_3_r` = 6 - `PPT-A_3`,
    `PPT-A_4_r` = 6 - `PPT-A_4`
  )

# Compute composite scores as row means
pilot_means <- pilot_csv %>%
  mutate(
    score_ppt_a    = rowMeans(across(c(`PPT-A_1`, `PPT-A_2`, `PPT-A_3_r`,
                                       `PPT-A_4_r`, `PPT-A_5`, `PPT-A_6`)), na.rm = TRUE),
    score_trust    = rowMeans(across(all_of(cols_trust)),    na.rm = TRUE),
    score_distrust = rowMeans(across(all_of(cols_distrust)), na.rm = TRUE),
    score_company  = rowMeans(across(all_of(cols_company)),  na.rm = TRUE)
  )

cor_data <- pilot_means %>%
  select(score_ppt_a, score_trust, score_distrust, score_company) %>%
  drop_na()

cor_matrix <- cor(cor_data, use = "complete.obs")
cor_matrix
```

```{r}
# Significance tests for each PTT-A correlation
cor.test(cor_data$score_ppt_a, cor_data$score_trust)
cor.test(cor_data$score_ppt_a, cor_data$score_distrust)
cor.test(cor_data$score_ppt_a, cor_data$score_company)
```

When interpreting correlations between PTT-A and the TPSai subscales, it is important to note that we do not assume that trust and distrust are opposite poles of a single continuum. Rather, consistent with two-dimensional conceptualisations of trust @mcknight2001 , distrust may constitute a psychologically distinct construct that can coexist with trust toward the same agent. However, the PTT-A scale conceptualises trust as a single dimension comprising trust and distrust @scholz2025.

Accordingly, while PTT-A is expected to correlate positively with the *Trust* subscale, its relationships with the *Distrust* and *Corporate Distrust* subscales may be weaker or asymmetric. Such a pattern would suggest that the TPSai scales capture distinct attitudinal dimensions toward generative AI for psychological support, rather than merely reflecting general propensity to trust automated systems.

If substantial correlations are observed, we will explore subtracting PTT-A from individual trust ratings to adjust for baseline trust tendencies. This adjustment would shift the trust rating scale from \[1, 5\] to a centred domain (e.g., \[-5, +5\]), isolating context-specific trust toward generative AI. In this scenario, analyses will be reported both with and without PTT-A adjustment to ensure transparency and assess the impact of baseline trust on factor structure, reliability, and validity.
